\chapter{Introduction}\label{chap:intro}

\setlength{\parindent}{1em}

\subsection{Motivation}

The fundamental virtue of a mathematical proof is that it provides certainty: a deductive argument shows that the assumptions of a mathematical statement logically guarantee the conclusion. In practice, however, informal, natural-language proofs are prone to imprecision, ambiguity, and error. Using a formal language such as Lean~4 \cite{LeanTheoremProver} removes such ambiguity and imprecision, enabling a proof assistant to verify correctness down to the primitives of a formal axiomatic system.

Formal proof assistants such as Lean, Rocq \cite{RocqTheoremProver}, and Isabelle \cite{IsabelleTheoremProver} have transformed mathematical practice by making correctness explicit and mechanized. Community libraries such as Lean's Mathlib \cite{Mathlib} now grow at a pace driven both by increasing human contributions and by recent advances in neural theorem proving and autoformalization \cite{achim2025aristotleimolevelautomatedtheorem, alphaproof}. This rapid expansion raises multiple concerns about data quality.

First, this expansion stresses the maintainability, coherence, and long-term usability of libraries due to proofs that are often heterogeneous in style and clarity. Instructors show their students how to shorten their proofs and structure them better, and the maintainers of Mathlib demand revisions to submissions to improve their robustness and adhere to style guidelines. Second, low library quality decreases its utility as training data: modern theorem provers and autoformalizers increasingly train on these very corpora, so the structure and readability of proofs directly shape downstream prover performance \cite{gu2025proofoptimizer}.

Although any two correct formal proofs of a statement equally establish the validity of their conclusion, there are various criteria on which one may be preferred over another. A proof may be more readable, more concise, more modular, or less dependent on obscure library lemmas. When an expert formalizer finishes a proof, they always go back and revise it --- aiming, for example, to improve readability and robustness. Unfortunately, the growth rate of formal libraries already exceeds what human reviewers can reliably curate, and this discrepancy is only expected to widen due to increasing volumes of machine-generated proofs, which even when correct do not carry similar guarantees as to their quality or understandability \cite{SeedProver}.

Moreover, automated proof optimization is not only useful in its own right, but also for improving AI that can find proofs on its own. At the very least, it provides a form of data augmentation: the limited amount of formal training data is a bottleneck for machine learning, and our methods provide ways of generating additional data automatically. Moreover, our methods also provide a means of optimizing the quality and structure of said training data for an arbitrary metric. 

For example, many modern theorem provers (e.g. \cite{alphaproof}, \cite{GodelProver}) rely on the draft-sketch-prove paradigm \cite{jiang2023draft}, which generates proofs by first sketching high level proof outlines with ``holes'' for subproofs/lemmas, and then recursively solving these lemmas. For that purpose, it is useful to have a corpus of proofs written in a structured, modular form that maximizes the number and balance of these subproofs. 

\subsection{Contributions}

We make the following contributions:
\begin{itemize}
    \item \textbf{A formalization of the proof optimization task.} We define proof optimization over a verified proof $(c,x,y_0)$ and an arbitrary user-specified metric $\mu$, separating the problem of finding a proof (neural theorem proving) from the problem of rewriting a proof to be shorter, more structured, or less dependent on external lemmas. We show that neural theorem proving is a strict special case of this framework.
    \item \textbf{ImProver: an agentic system for proof optimization.} We introduce ImProver (Chapters~\ref{chap:improver}--\ref{chap:improver-exp}), a GPT-4o-based agent that rewrites Lean proofs, outperforms a naive GPT-4o baseline by $423$--$778\%$ across undergraduate-level proofs.
    \item \textbf{ImProver~2: a trained small model for proof optimization.} We introduce ImProver~2 (Chapters~\ref{chap:improver2}--\ref{chap:improver2-exp}), a pipeline that bootstraps a 7B-parameter model to optimize proofs using a novel IRPO RL pipeline and upgraded neurosymbolic scaffold, outperforming models orders of magnitude larger and extends ImProver to research-level proof optimization.
    \item \textbf{Empirical evidence that formal-structure exposure is the key lever.} Across both systems, gains track neurosymbolic augmentation rather than model scale alone: the scaffold nearly doubles length improvement on DeepSeek-R1 7B before any preference optimization, and also boosts frontier models when applied at inference time. We argue this supports the thesis that proof optimization is primarily a specialization problem, not a scale problem.
\end{itemize}

\subsection{Thesis Overview}

Motivated by the need for more efficient and effective proof optimization, this thesis presents the ImProver and ImProver 2 systems, first exploring a simple yet effective agentic scaffolding approach, and then generalizing to a more powerful bootstrapped RL training pipeline.

ImProver (Chapters~\ref{chap:improver}--\ref{chap:improver-exp}) is the first system for automated proof optimization. It is a large language model agent that rewrites Lean proofs to optimize arbitrary user-defined metrics, incorporating Chain-of-States (CoS) prompting, iterative refinement, and retrieval-augmented generation (RAG). ImProver is capable of rewriting undergraduate, competition, and research-level proofs so that they are substantially shorter and more declarative, outperforming naive GPT-4o by $423$--$778\%$ depending on the metric. It also demonstrates that proof optimization generalizes neural theorem proving.

ImProver~2 (Chapters~\ref{chap:improver2}--\ref{chap:improver2-exp}) addresses the core limitation of ImProver: its reliance on expensive, proprietary large language models. Namely, ImProver~2 trains a 7B-parameter small language model (SLM) to optimize Lean proofs via an IRPO-based iterative self-improvement pipeline, as well as extending the agentic scaffolding techniques of ImProver to apply to more complex metrics and research-level datasets. The resulting model outperforms models orders-of-magnitude larger across metrics, and is even competitive with frontier models such as GPT-5-high on particular metrics.

\subsection{Publications.} 

The material in Chapters~\ref{chap:improver} and \ref{chap:improver-exp} is based on the paper ``ImProver: Agent-Based Automated Proof Optimization'' \cite{ImProver}, published at ICLR 2025. The material in Chapters~\ref{chap:improver2} and \ref{chap:improver2-exp} is based on the paper ``ImProver~2: Iteratively Self-Improving LMs for Neurosymbolic Proof Optimization'' \cite{ImProver2}, submitted to ICML 2026.
