\chapter{Centralized Multi-Agent Theorem Proving}\label{chap:centralized}

\newcommand{\proves}{\vdash}




With the motivation of modelling real-world mathematical development in mind, we first introduce a formalism of the core structure of mathematical tasks that will be used throughout this thesis. Inspired by proof-theoretic and type-theoretic semantics, we represent mathematical statements as formulas within a formal language, and the process of proving these statements as the resolution of proof obligations (sequents) that are defined by these formulas. We then introduce a library structure that models the evolving state of mathematical knowledge across multiple agents, each with private information.

Then with this formalism in place, we introduce the centralized multi-agent theorem proving problem, a POMDP in which a single centralized planner allocates compute across proof/conjecture/publication actions for multiple agents with private information. We study this problem under the simplified environment of perfect public information, and study its complexity and approximation-theoretic properties by relating it to stochastic online scheduling problems with precedence constraints. We show that the problem is intractable in general, but identify special cases that are tractable and/or admit good approximations. We also discuss the limitations of this centralized formulation in the full environment of private information, and motivate the need for a decentralized formulation in which coordination is mediated by a market mechanism, which we study in the next chapter.


\subsection{Background}

\subsubsection{Syntax}

We consider the \textit{formula} $\phi$ as the core syntactic primitive of (formal) mathematics. Namely, we represent mathematical theorem statements syntactically as a formula (i.e. object of type \texttt{Prop}) within some formal language $\Sigma$, such as Lean, Rocq, etc. We define $\mathcal{F}$ as the collection of all such formulas, and equip it with an involution $\lnot : \mathcal{F}\rightarrow \mathcal{F}$; namely, $\lnot(\lnot\varphi)=\varphi$ for all $\varphi\in\mathcal{F}$. Note that we will refer to formulas as ``nodes'' interchangably due to a natural graphical structure that will be discussed later.

\subsubsection{Semantics}

We now aim to represent the semantic meaning of a theorem statement within a particular context concisely. Towards this, we consider the \textit{sequent}, given as an element of $\mathcal{S} := \mathcal{P}_{fin}(\mathcal{F}) \times \mathcal{F} $ and stylized as $[\Gamma \proves \phi]$ (read: ``Gamma \textit{proves} phi''). Namely, such a sequent represents a \textit{proof obligation}, i.e. the notion of proving a theorem statement $\phi$ (the \textit{consequent}) within the \textit{antecedent} context $\Gamma$. Such a proof obligation may be \textit{resolved}, meaning proven.

\subsubsection{Libraries}

We assume the existence of a universal clock $t \in \mathbb N$ and set of agents $\mathcal N$, and use it to define a library $\mathcal{L}_\bullet = \{\mathcal{L}_t\}_{t \in \mathbb{N}}$ as a collection of discrete library states that evolves over time. At every discrete timestep $t \in \mathbb{N}$, each library state is given as a tuple 
\[\mathcal{L}_t := (\mathcal{C}_t,\mathcal{RS}_t, \delta_t, \Theta_t) \in \mathscr{L}\]
where we partition $\mathcal{F} = \mathcal{C}_t \cup \mathcal{G}_t, \mathcal{C}_t \cap \mathcal{G}_t = \varnothing$ into \textit{concrete} and \textit{ghost} nodes. Intuitively, concrete nodes represent theorem statements that have been stated or conjectured, and ghost nodes represent all hypothetical, possible theorem statements that are yet to be conjectured or stated. $\mathcal{C}_t$ naturally induces a semantic analogue: let $\mathcal{CS}_t := \mathcal{P}_{fin}(\mathcal{C}_t) \times \mathcal{C}_t$ be the set of all \textit{concrete sequents}, which contains a subset $\mathcal{RS}_t \subseteq \mathcal{CS}_t$ of \textit{resolved sequents}, which represent the concrete nodes that have been proven (within a particular antecedent context, of course). 


Mathematically, a proof of a particular theorem in some context often explicitly depends on other lemmas within this context. In an effort to model this phenomenon, define $\delta_t : \mathcal{RS}_t \to \mathcal{P}_{fin}(\mathcal{F})$ as the dependency mapping.
Note that in mathematics, there may exist many proofs of the same theorem with different dependencies for each, and one may find new proofs of the same theorem that are also valid. Thus, in the most general formulation, we only require that resolved sequents must remain resolved ($\mathcal{RS}_{t} \subseteq \mathcal{RS}_{t+1}$), though the underlying ``proof'' (moreso, the set of dependencies of this ``proof'') can change.

And lastly, we define $\Theta_t : \mathcal{RS}_t\to\mathbb N^+ \times \mathcal N$ to record the (discrete) solve-time associated to each resolved sequent, and the agent responsible for the resolution. 


So, in other words, the library state $\mathcal{L}_t$ represents the collection of all stated theorem statements, the collection of proof witnesses within particular contexts for a subset of these statements, and the dependencies and ancillary metadata of these witnesses.

\subsubsection{Library Invariants and Properties}

Naturally, we impose a monotonicity invariant for libraries: one may not ``unconjecture'' or ``unprove'' a theorem. Formally, given a library $\mathcal{L}_\bullet = \{\mathcal{L}_t\}_{t \in \mathbb{N}}=\{(\mathcal{C}_t,\mathcal{RS}_t,\delta_t,\Theta_t)\}_{t \in \mathbb{N}}$, we require that for all $t \in \mathbb{N}$:
\[\mathcal{C}_{t} \subseteq \mathcal{C}_{t+1} \quad \text{and} \quad \mathcal{RS}_{t} \subseteq \mathcal{RS}_{t+1}\]

Additionally, we require that if one conjectures a theorem, they must also have automatically conjectured its converse; in other words:
\[\phi \in \mathcal{C}_t \iff \neg\phi \in \mathcal{C}_t\]

Similarly, one may not prove both a theorem and its converse true: 
\[[\Gamma \proves \phi] \in \mathcal{RS}_t \implies [\Gamma \proves \neg \phi] \notin \mathcal{RS}_t\]

Semantically, we note reflexivity holds: $\phi \in \Gamma \implies [\Gamma \proves \phi] \in \mathcal{RS}_t$. And for the sake of consistency, we require that for any $[\Gamma \proves \phi] \in \mathcal{CS}_t$, $\psi \in \Gamma \implies \neg\psi \notin \Gamma$ (although we do not impose any full propositional semantics or richer logical theory behind these abstract formulas, preferring to leave them fully symbolic objects with various resolution and concreteness states, allowing inconsistent antecedents in our witnesses limits the scalability of our approach to deployment as a strategy module for multiagent theorem proving in a fully-defined formal language and type theory). 

Note that this last condition implicitly redefines
\[\mathcal{CS}_t := \mathcal{P}_{fin}(\mathcal{C}_t)/\sim \times \mathcal{C}_t \] where $\phi\sim\neg\phi$ is the equivalence relation generated by the negation operation. So, due to this consistency condition, we note that $\mathcal{RS}_t$ is upwards-closed, in that if $[\Gamma \proves \phi] \in \mathcal{RS}_t$, then for all $\Gamma' \in \mathcal{P}_{fin}(\mathcal{C}_t)$, $\Gamma \subseteq \Gamma' \implies [\Gamma'\proves\phi]\in \mathcal{RS}_t$.

Additionally, considering the dependency mapping, we note that intuitively $\delta_t$ must satisfy the property that $\delta_t([\Gamma \proves \phi]) \subseteq \Gamma$. Moreover, one may use $\delta_t$ to induce a directed graph $D(\delta_t) := (\mathcal{CS}_t,\mathcal{E}_t)$, where 
\[\mathcal{E}_t = \Big\{([\Gamma\proves \phi], [\Gamma' \proves \psi]) : \phi \in \delta_t([\Gamma' \proves \psi]), \Gamma\subseteq \Gamma'\Big\}\]
We require that this induced directed graph $D(\delta_t)$ of $\delta_t$ must be acyclic.

\subsubsection{Axioms and Fully Resolved Sequents}

As an aside, we additionally note that although $[\Gamma \proves \phi] \in \mathcal{RS}_t$ may be marked resolved, that does not mean that $\phi$ has a ``error''-free proof within the current library state as its proof may depend on some still-unresolved lemmas. More concretely, we may define $\mathcal{FRS}_t \subseteq \mathcal{RS}_t$ as the collection of \textit{fully resolved sequents}, which is defined inductively such that \textit{axioms} $[\proves \phi] \in \mathcal{FRS}_t$ and $[\Gamma \proves \phi] \in \mathcal{FRS}_t$ if and only if for all $\phi' \in \delta_t([\Gamma \proves \phi])$, there exists $\Gamma'\subseteq \Gamma$ such that $[\Gamma' \proves \phi'] \in \mathcal{FRS}_t$.


\subsubsection{Private and Shared Libraries}
In the multiagent setting, each agent $\alpha\in\mathcal{N}$ maintains a \textit{private library} $\mathcal{L}_\bullet^\alpha = \{\mathcal{L}^\alpha_t\}_{t\in\mathbb{N}}$ with
\[
\mathcal{L}^\alpha_t := (\mathcal{C}^\alpha_t,\mathcal{RS}^\alpha_t,\delta^\alpha_t,\Theta_t^\alpha),
\]
and the environment maintains a \textit{shared (public) library} $\mathcal{L}_\bullet^0 = \{\mathcal{L}^0_t\}_{t\in\mathbb{N}}$ with
\[
\mathcal{L}^0_t := (\mathcal{C}^0_t,\mathcal{RS}^0_t,\delta^0_t, \Theta_t^0).
\]

We additionally enforce the following inclusion invariant for all $t \in \mathbb{N}, \alpha \in \mathcal{N}$:
\[
\mathcal{L}^0_t \subseteq \mathcal{L}^\alpha_t
\]
meaning $\mathcal{C}^0_t\subseteq \mathcal{C}^\alpha_t$, $\mathcal{RS}^0_t\subseteq \mathcal{RS}^\alpha_t$, and for all $s\in \mathcal{RS}_t^0, \delta^0_t(s) = \delta^\alpha_t(s)$ and $\Theta^0_t(s)=\Theta^\alpha_t(s)$.


\subsection{The Centralized MATP Problem}

We now introduce the Centralized Multi-Agent Theorem Proving (MATP) problem, a POMDP in which a single centralized planner allocates compute across proof/conjecture/publication actions for multiple agents with private information.

\subsubsection{Background: POMDP}
We first recall the standard Markov Decision Process (MDP) formalism, which models a single agent sequential decision making problem with full state observability.

\begin{definition}[Markov Decision Process]
A discrete-time Markov decision process is a tuple
\[
\mathcal{M} = (\mathsf{S},\;\mathsf{A},\;T,\;T^0,\;c),
\]
where $\mathsf{S}$ is a measurable state space, $\mathsf{A}$ is a measurable action space, $T(\cdot\mid s,a)\in\mathcal{P}(\mathsf{S})$ is a transition kernel, $T^0\in\mathcal{P}(\mathsf{S})$ is an initial state distribution, and $c:\mathsf{S}\times\mathsf{A}\to\mathbb{R}$ is a per-step cost function. At time $t=0$, nature draws $s_0\sim T^0$. At each subsequent time $t\ge 0$, the controller selects $a_t$, incurs cost $c(s_t,a_t)$, and the environment transitions to $s_{t+1}\sim T(\cdot\mid s_t,a_t)$.
\end{definition}

One may generalize the MDP formalism to include partial observability, which is more appropriate for modelling the multiagent theorem proving problem in its full generality, as each agent has private information in the form of their private library state. The natural formulation of this problem is as a Partially Observable Markov Decision Process (POMDP), in which the agent does not directly observe the underlying state, but rather receives observations that are generated from the underlying state according to some observation function.

\begin{definition}[POMDP]
A discrete-time partially observable Markov decision process is a tuple
\[
\mathcal{P} = (\mathsf{S},\; \mathsf{A},\; \mathsf{O},\; T,\; T^0,\; Z,\; c),
\]
for a state space $\mathsf{S}$, a measurable space whose elements encode the full (hidden) state of the environment; an action space $\mathsf{A}$, a measurable space of actions available to the controller; an observation space $\mathsf{O}$, a measurable space of observations received by the controller; a transition kernel $T(\cdot \mid s, a) \in \mathcal{P}(\mathsf{S})$ specifying the distribution of the next state given a current state and action; an initial state distribution $T^0 \in \mathcal{P}(\mathsf{S})$; an observation kernel $Z(\cdot \mid s) \in \mathcal{P}(\mathsf{O})$ specifying the distribution of the observation received by the controller when the environment is in state $s$; and a per-step cost function $c : \mathsf{S} \times \mathsf{A} \to \mathbb{R}$.
\end{definition}
 
The process evolves as follows. At time $t = 0$, nature draws a latent state
$s_0 \sim T^0$ and the controller receives an observation $o_0 \sim Z(\cdot \mid s_0)$.
At each subsequent time $t \ge 0$, the controller selects an action $a_t$, the environment
transitions to a new state $s_{t+1} \sim T(\cdot \mid s_t, a_t)$, and the controller receives
a new observation $o_{t+1} \sim Z(\cdot \mid s_{t+1})$.
 
The controller's observation history at time $t$ is the tuple
\[
h_t := (o_0,\, a_0,\, o_1,\, a_1,\, \ldots,\, a_{t-1},\, o_t),
\]
which constitutes the controller's complete information record. 
 
\begin{definition}[Policy]
A randomized nonanticipative policy is a family of stochastic kernels
$\pi = (\pi_t)_{t \ge 0}$ with $\pi_t : \mathsf{H}_t \to \mathcal{P}(\mathsf{A})$, where $\mathsf{H}_t := (\mathsf{O}\times\mathsf{A})^t\times\mathsf{O}$ is the space of observation histories of length $t$, such that $a_t \sim \pi_t(\cdot \mid h_t)$ depends only on the observation history up to time $t$. Equivalently, the action process $(a_t)_{t \ge 0}$ is adapted to the observation filtration $\mathcal{F}_t^{\mathrm{obs}} := \sigma(o_0, a_0, \ldots, a_{t-1}, o_t)$.
\end{definition}
 
\begin{remark}[Belief state]
By standard POMDP theory, the posterior belief
$b_t := \mathrm{Law}(s_t \mid h_t) \in \mathcal{P}(\mathsf{S})$ is a sufficient statistic of
the observation history for optimal control: there exists an optimal policy whose action
at time $t$ depends on $h_t$ only through $b_t$. One may equivalently reformulate the
POMDP as a fully observed MDP on the belief space $\mathcal{P}(\mathsf{S})$, the \emph{belief MDP}, in which the state is $b_t$ and the transition is given by Bayesian updating.
\end{remark}

To solve a POMDP means to find a policy $\pi$ that minimizes the expected total (discounted) cost:
\[
J(\pi) := \mathbb{E}^{\pi}\!\left[\sum_{t=0}^{\infty}\gamma^t\, c(s_t,a_t)\;\middle|\; s_0\sim T^0\right]
\]
for a discount factor $\gamma\in[0,1)$, or the expected total cost under a finite horizon $H$:
\[
J_H(\pi):=\mathbb{E}^\pi\!\left[\sum_{t=0}^{H-1}c(s_t,a_t)\;\middle|\;s_0\sim T^0\right].
\]

\subsubsection{Background: Action Primitives}
We now specify the action primitives that the centralized planner may assign to individual agents. These primitives are standalone kernels and update maps that will be assembled into the POMDP transition structure.
 
\paragraph{Discrete kernels.}
All stochastic primitives in this section have finite or countable output spaces, and may be treated as
probability mass functions (pmfs). Concretely, for an input space $X$ and discrete output space $Y$,
a stochastic kernel is a map $K:Y\times X\to[0,1]$ such that $\sum_{y\in Y}K(y\mid x)=1$ for all $x\in X$.
 
\paragraph{Proof Kernel.}
 
Fix an agent $\alpha\in\mathcal{N}$. The proof kernel models allocating a time budget to attempt to resolve a target sequent. Let the input space be
\[
X_{\mathrm{proof}} := \mathscr{L}\times\mathcal{S}\times \mathbb{N}^+,
\]
where the components are a library state, a target sequent $s=[\Gamma\proves\phi]$, and a discrete time budget $\tau$. Let the output space be
\[
Y_{\mathrm{proof}} := \{(0,\bot)\}\;\cup\;\{(1,d): d\in\mathcal{P}_{\mathrm{fin}}(\mathcal{F})\},
\]
where $(0,\bot)$ denotes failure, and $(1,d)$ denotes success together with a dependency set $d$. The proof kernel for agent $\alpha$ is a pmf
\[\mathcal{K}^{\alpha}_{\mathrm{proof}} : Y_{\mathrm{proof}}\times X_{\mathrm{proof}} \to [0,1].\]
On admissible inputs $(\mathcal{L},[\Gamma\proves\phi],\tau)$, a successful output $(1,d)$ satisfies $d\subseteq\Gamma$.
 
\paragraph{Conjecture Kernel.}
 
Fix an agent $\alpha\in\mathcal{N}$. The conjecture kernel models allocating a time budget to propose a new formula, anchored at an existing sequent. Let the input space be
\[
X_{\mathrm{conj}} :=\mathscr{L}\times \mathcal{S}\times \mathbb{N}^+,
\]
and the output space be
\[
Y_{\mathrm{conj}} := \{(0,\bot)\}\;\cup\;\{(1,\psi): \psi\in\mathcal{F}\}.
\]
The conjecture kernel for agent $\alpha$ is a pmf
$\mathcal{K}^{\alpha}_{\mathrm{conj}} : Y_{\mathrm{conj}}\times X_{\mathrm{conj}} \to [0,1]$.
Given admissible input $(\mathcal{L}, [\Gamma \proves \phi], \tau)$ with $\mathcal{L} = (\mathcal{C},\mathcal{RS},\delta,\Theta)$ and a successful output $(1,\psi)$, we enforce $\psi \in \mathcal{F} \setminus \mathcal{C}$.
 
\paragraph{Dependency Closure for Publication.}
 
Publication is a deterministic primitive that copies a closure-completed subset of an agent's private library
into the shared library. Fix agent $\alpha$ and time $t$, and suppose $\delta_t^\alpha$ induces the DAG
$D(\delta_t^\alpha)=(\mathcal{CS}_t^\alpha,\mathcal{E}_t^\alpha)$. For any set of target resolved sequents $R\subseteq \mathcal{RS}_t^\alpha$, define its \emph{dependency closure}:
\begin{align*}
\mathrm{cl}^{\alpha}_t(R) := 
\Big\{\, &s\in \mathcal{RS}_t^\alpha : \exists\, r\in R \text{ such that there}\\& \text{is a directed path } s\to r
\text{ in } D(\delta_t^\alpha)\,\Big\}
\end{align*}
 
A publication payload is a pair $U_t^\alpha=(C,R)$ with $C\subseteq \mathcal{C}_t^\alpha,\; R\subseteq \mathcal{RS}_t^\alpha$. Its closure-completion is $\overline{U_t^\alpha}:=(\overline{C},\overline{R})$ where $\overline{R}:=\mathrm{cl}^\alpha_t(R)$ and
\[\overline{C}:= C \;\cup\; \bigcup_{[\Gamma\proves\phi]\in \overline{R}}(\Gamma\cup\{\phi\}).\]
 
\paragraph{Query Model.}
 
Each agent $\alpha$ maintains a stateful query model $Q_t^\alpha \in \mathcal{Q}$ that estimates prover performance. A query readout is a map
$q^\alpha : \mathcal{Q}\times \mathcal{S} \to [0,1]\times \mathbb{N}$, returning $(\hat p_t^\alpha(s),\hat \tau_t^\alpha(s)) := q^\alpha(Q_t^\alpha,s)$, where $\hat p_t^\alpha(s)$ estimates the probability that sequent $s$ is resolvable by agent $\alpha$, and $\hat\tau_t^\alpha(s)$ estimates the expected time-to-solution.
 
The query model is updated via an operator $\Delta^\alpha : \mathcal{Q}\times \mathcal{D}(\mathcal{Q}) \to \mathcal{Q}$, where $\mathcal{D}(\mathcal{Q}) := \mathcal{S} \times (\mathbb{N}^+ \times \mathcal{N})$ is the space of query-training data. We enforce commutativity and associativity of $\Delta^\alpha$ so that batch updates are well-defined and order-independent.
 
 
\subsubsection{Definition}
 
Fix a set of agents $\mathcal{N}$ and a target set of sequents $\Phi^*\subseteq \mathcal{S}$. The \emph{Centralized Multi-Agent Theorem Proving} (MATP) problem is specified by the POMDP
\[
\mathcal{P}_{\mathrm{MATP}} = (\mathsf{S},\;\mathsf{A},\;\mathsf{O},\;T,\;T^0,\;Z,\;c)
\]
defined as follows.
 
 
\paragraph{State Space $\mathsf{S}$.}
 
A state at time $t$ is a tuple
\begin{align*}
s_t :=
\Big(
&\{\mathcal{L}^\alpha_t\}_{\alpha\in\mathcal{N}},\;
\mathcal{L}^0_t,\;
\{J^\alpha_t\}_{\alpha\in\mathcal{N}},\;
\{Q^\alpha_t\}_{\alpha\in\mathcal{N}},
\\&
\{\Omega^\alpha_t\}_{\alpha\in\mathcal{N}},\;
\{U_{t}^\alpha\}_{\alpha\in\mathcal{N}}
\Big)\in\mathsf{S},
\end{align*}
where:
\begin{itemize}
\item $\mathcal{L}^\alpha_t\in\mathscr{L}$ is agent $\alpha$'s private library and
$\mathcal{L}^0_t\in\mathscr{L}$ is the shared public library, subject to the inclusion invariant $\mathcal{L}^0_t \subseteq \mathcal{L}^\alpha_t$ for all $\alpha\in\mathcal{N}$.
\item $J^\alpha_t$ is the \emph{durative job descriptor}:
\[
J^\alpha_t \in \{\bot\}\ \cup\ \big(\{\mathrm{Proof}, \mathrm{Conj}\}\times\mathcal{S}\times\mathbb{N}^+\times\mathbb{N}^+\big),
\]
with $\bot$ meaning idle and $(\cdot,s,\tau_{\mathrm{rem}},\tau_{\mathrm{eff}})$ encoding a job of type $\mathrm{Proof}$ or $\mathrm{Conj}$ on target sequent $s$ with $\tau_{\mathrm{rem}}$ steps remaining in the current run and $\tau_{\mathrm{eff}}$ the effective cumulative budget used to evaluate the kernel upon completion.
\item $Q^\alpha_t\in\mathcal{Q}$ is agent $\alpha$'s query model state.
\item $\Omega^\alpha_t \in \{\bot\}\ \sqcup\ (\mathcal{S}\times[0,1]\times\mathbb{N})$ is agent $\alpha$'s last query response, either empty $\bot$ or a triple $(s,\hat p,\hat \tau)$.
\item $U^\alpha_{t} : \mathcal{S} \times \{\mathrm{proof},\mathrm{conj}\} \to \mathbb{N}$ records the total time already spent by agent $\alpha$ on proof or conjecture attempts for each sequent, with $U^\alpha_0 \equiv 0$.
\end{itemize}
We take $\mathsf{S}$ to be the set of all such tuples satisfying the constituent objects' invariants.
 
 
\paragraph{Action Space $\mathsf{A}$ and Admissibility.}
 
The centralized planner selects a \emph{joint action} $a_t = (a^\alpha_t)_{\alpha\in\mathcal{N}} \in \mathsf{A}$, where $\mathsf{A}:= \prod_{\alpha\in\mathcal{N}}\mathcal{A}^\alpha$ and each per-agent action space is
\begin{align*}
\mathcal{A}^\alpha
:=\;&
\{\mathrm{Prove}\}\times(\mathcal{S}\times\mathbb{N}^+)
\\&\sqcup\;
\{\mathrm{Conj}\}\times(\mathcal{S}\times\mathbb{N}^+)
\\&\sqcup\;
\{\mathrm{Pub}\}\times\big(\mathcal{P}_{\mathrm{fin}}(\mathcal{F})\times \mathcal{P}_{\mathrm{fin}}(\mathcal{S})\big)
\\&\sqcup\;
\{\mathrm{Qry}\}\times \mathcal{S}
\\&\sqcup\;\{\bot\}.
\end{align*}
 
Define the admissibility predicate $\mathrm{Adm}:\mathsf{S}\times\mathsf{A}\to\{0,1\}$ componentwise: a joint action $a=(a^\alpha)_{\alpha\in\mathcal{N}}$ is admissible at state $s$ if and only if $\mathrm{Adm}^\alpha(s,a^\alpha)=1$ for all $\alpha\in\mathcal{N}$, where:
\begin{itemize}
    \item If $J^\alpha_t\neq\bot$, the agent is occupied and forced to no-op:
    \[
    \mathrm{Adm}^\alpha(s_t,a^\alpha)=\mathbf{1}[a^\alpha=\bot].
    \]
    \item If $J^\alpha_t=\bot$, the per-action admissibility conditions are:
    \begin{itemize}
    \item $\mathrm{Adm}^\alpha(s_t,\bot)=1$ always.
    \item For $a^\alpha = (\mathrm{Prove},(s,\tau))$ or $(\mathrm{Conj},(s,\tau))$:
    $\mathrm{Adm}^\alpha(s_t,a^\alpha)=1 \iff s \in \mathcal{CS}_t^\alpha.$
    \item For $a^\alpha = (\mathrm{Pub},(C,R))$: $\mathrm{Adm}^\alpha(s_t,a^\alpha)=1 \iff C\subseteq\mathcal{C}_t^\alpha$ and $R\subseteq\mathcal{RS}_t^\alpha.$
    \item For $a^\alpha = (\mathrm{Qry},s)$: $\mathrm{Adm}^\alpha(s_t,a^\alpha)=1 \iff s \in \mathcal{CS}_t^\alpha.$
    \end{itemize}
\end{itemize}
 
\begin{remark}[Centralization and admissibility]
\label{rem:centralized-admissibility}
A crucial subtlety: the admissibility predicate for agent $\alpha$ depends on $\alpha$'s private library $\mathcal{L}^\alpha_t$, which the centralized planner may not fully observe. Thus, the planner must reason about which actions are admissible for each agent based on its belief over the hidden state. This is a fundamental source of difficulty in the centralized formulation with private information, and is precisely what motivates the connection to mechanism design and the Nisan--Ronen conjecture in Section~\ref{sec:limits-centralization}.
\end{remark}
 
 
\paragraph{Observation Space $\mathsf{O}$ and Observation Kernel $Z$.}
 
The centralized planner observes the public projection of the state:
\[
o_t :=
\big(\mathcal{L}^0_t,\;\{\Omega^\alpha_t\}_{\alpha\in\mathcal{N}}\big)
\in\mathsf{O}.
\]
That is, the planner observes the shared library and the most recent query response from each agent. In particular, the private libraries $\{\mathcal{L}^\alpha_t\setminus\mathcal{L}^0_t\}_{\alpha\in\mathcal{N}}$, internal job states, and query model parameters are \emph{hidden}.
 
The observation kernel is deterministic:
\[
Z(o\mid s)=\mathbf{1}\!\big[\,o=\mathrm{Obs}(s)\,\big],
\]
where $\mathrm{Obs}:\mathsf{S}\to\mathsf{O}$ computes the public projection as above.
 
\begin{remark}[Perfect public information]
\label{rem:perfect-public-info}
In the special case where all libraries are public---i.e., $\mathcal{L}^\alpha_t = \mathcal{L}^0_t$ for all $\alpha\in\mathcal{N}$ and all $t$---the observation $o_t$ determines $s_t$ and the POMDP reduces to a fully observed MDP. We refer to this as the \emph{perfect public information} (PPI) setting and study it in Sections~\ref{sec:scheduling} and~\ref{sec:complexity}.
\end{remark}
 
 
\paragraph{Transition Kernel $T$.}
 
We define $T(\cdot\mid s_t,a_t)$ by a three-step generative construction: a deterministic pre-update, a stochastic job-completion draw, and a deterministic post-update.
 
\subparagraph{Step 1: Deterministic Pre-Update.}
 
Given a state $s_t\in\mathsf{S}$ and admissible joint action $a_t\in\mathsf{A}$, define
$\tilde s_t := \mathrm{Pre}(s_t,a_t)$ by applying the following edits in order:
\begin{enumerate}
    \item \textbf{Decrement job timers.} For each $\alpha \in \mathcal{N}$: if $J^\alpha_t=\bot$, leave it unchanged; if $J^\alpha_t=(\cdot,s,\tau_{\mathrm{rem}},\tau_{\mathrm{eff}})$ with $\tau_{\mathrm{rem}}>0$, set
$\tau_{\mathrm{rem}}\leftarrow \tau_{\mathrm{rem}}-1$.
 
\item \textbf{Start new jobs.} For each idle agent $\alpha$ (i.e., $J^\alpha_t=\bot$):
\begin{itemize}
    \item If $a^\alpha_t=(\mathrm{Prove},(s,\tau))$, set
$J^\alpha \leftarrow (\mathrm{Proof},s,\tau,\tau+U^\alpha_t(s,\mathrm{proof}))$.
	\item If $a^\alpha_t=(\mathrm{Conj},(s,\tau))$, set
$J^\alpha \leftarrow (\mathrm{Conj},s,\tau,\tau+U^\alpha_t(s,\mathrm{conj}))$.
\end{itemize}
 
\item \textbf{Apply publications.} For each $\alpha\in\mathcal{N}$ with $a^\alpha_t=(\mathrm{Pub},(C^\alpha,R^\alpha))$: compute the closure-completion $\overline{(C^\alpha,R^\alpha)}=(\overline{C}^\alpha,\overline{R}^\alpha)$ and update:
\[\mathcal C^0 \leftarrow \mathcal C^0 \cup \textstyle\bigcup_{\alpha}\overline C^\alpha,
\qquad
\mathcal{RS}^0 \leftarrow \mathcal{RS}^0 \cup \textstyle\bigcup_{\alpha}\overline R^\alpha.\]
For each newly resolved $s \in \mathcal{RS}^0_{\mathrm{new}} := \mathcal{RS}^0 \setminus \mathcal{RS}^0_{\mathrm{old}}$, set $\delta^0(s) \leftarrow \delta^\alpha(s)$ and $\Theta^0(s) \leftarrow \Theta^\alpha(s)$.
Propagate to each agent $\beta$:
\[\mathcal C^\beta \leftarrow \mathcal C^\beta \cup (\mathcal C^0\setminus \mathcal C^0_{\mathrm{old}}),\quad
\mathcal{RS}^\beta \leftarrow \mathcal{RS}^\beta \cup \mathcal{RS}^0_{\mathrm{new}},\]
and for each $s\in \mathcal{RS}^0_{\mathrm{new}}$:
$\delta^\beta(s)\leftarrow\delta^0(s)$, $\Theta^\beta(s)\leftarrow\Theta^0(s)$, and $Q^\beta \leftarrow \Delta^\beta(Q^\beta,(s,\Theta^0(s)))$.
 
\item \textbf{Update query buffers.} For each $\alpha\in\mathcal{N}$: if $a^\alpha_t=(\mathrm{Qry},s)$, set
$\Omega^\alpha \leftarrow (s,\hat p,\max\{0,\hat\tau-U^\alpha_t(s,\mathrm{proof})\})$
where $(\hat p,\hat\tau):=q^\alpha(Q^\alpha,s)$; otherwise set $\Omega^\alpha\leftarrow \bot$.
\end{enumerate}
 
\subparagraph{Step 2: Stochastic Job Completion.}
 
Given $\tilde s_t$, define the sets of completing agents:
\[
    \mathsf{Comp}_{\mathrm{proof}}(\tilde s_t) = \{\alpha : J^\alpha(\tilde s_t)=(\mathrm{Proof},s,0,\tau)\text{ for some }s,\tau\},
\]
\[
\mathsf{Comp}_{\mathrm{conj}}(\tilde s_t) = \{\alpha : J^\alpha(\tilde s_t)=(\mathrm{Conj},s,0,\tau)\text{ for some }s,\tau\}.
\]
 
For each completing agent, we draw an independent outcome from the appropriate kernel. The joint outcome space is
\[Y(\tilde s_t)
:=
\prod_{\alpha\in\mathsf{Comp}_{\mathrm{proof}}(\tilde s_t)} Y_{\mathrm{proof}}
\;\times\;
\prod_{\alpha\in\mathsf{Comp}_{\mathrm{conj}}(\tilde s_t)} Y_{\mathrm{conj}},\]
equipped with the product measure
\begin{align*}
\mathbb K_{\tilde s_t}(dy)
&:=
\prod_{\alpha\in\mathsf{Comp}_{\mathrm{proof}}(\tilde s_t)}
\mathcal K^\alpha_{\mathrm{proof}}\!\big(dy^\alpha \mid \mathcal L^\alpha(\tilde s_t), s^\alpha_{\mathrm{proof}},\tau^\alpha_{\mathrm{proof}}\big)
\\&\quad\cdot
\prod_{\alpha\in\mathsf{Comp}_{\mathrm{conj}}(\tilde s_t)}
\mathcal K^\alpha_{\mathrm{conj}}\!\big(dy^\alpha \mid \mathcal L^\alpha(\tilde s_t), s^\alpha_{\mathrm{conj}},\tau^\alpha_{\mathrm{conj}}\big),
\end{align*}
where $(s^\alpha_{\cdot},\tau^\alpha_{\cdot})$ are read from $J^\alpha(\tilde s_t)$.
 
\subparagraph{Step 3: Deterministic Post-Update.}
 
Given $\tilde s_t$ and realized outcomes $y\in Y(\tilde s_t)$, define $s_{t+1} := \mathrm{Post}(\tilde s_t,y)$ by applying:
\begin{enumerate}
    \item \textbf{Resolve proof completions.} For each $\alpha\in\mathsf{Comp}_{\mathrm{proof}}(\tilde s_t)$ with outcome $y^\alpha = (b^\alpha, z^\alpha)$ and stored target/budget $(s^\alpha,\tau^\alpha)$: set $U^\alpha(s^\alpha,\mathrm{proof}) \leftarrow \tau^\alpha$. If $b^\alpha=1$ and $z^\alpha=d$:
\begin{itemize}
    \item $\mathcal{RS}^\alpha \leftarrow \mathcal{RS}^\alpha \cup \{s^\alpha\}$,
    \item $\delta^\alpha(s^\alpha) \leftarrow d$ \;(subject to $d\subseteq \Gamma$ and acyclicity),
    \item $\Theta^\alpha(s^\alpha)\leftarrow (\tau^\alpha,\alpha)$,
    \item $Q^\alpha \leftarrow \Delta^\alpha(Q^\alpha,(s^\alpha,\Theta^\alpha(s^\alpha)))$.
\end{itemize}
 
\item \textbf{Resolve conjecture completions.} For each $\alpha\in\mathsf{Comp}_{\mathrm{conj}}(\tilde s_t)$ with outcome $y^\alpha = (b^\alpha, z^\alpha)$: set $U^\alpha(s^\alpha,\mathrm{conj}) \leftarrow \tau^\alpha$. If $b^\alpha=1$ and $z^\alpha=\psi$:
$\mathcal C^\alpha \leftarrow \mathcal C^\alpha \cup \{\psi,\neg\psi\}$.
 
\item \textbf{Clear completed jobs.} For each $\alpha\in \mathsf{Comp}_{\mathrm{proof}}(\tilde s_t)\cup \mathsf{Comp}_{\mathrm{conj}}(\tilde s_t)$: set $J^\alpha \leftarrow \bot$.
\end{enumerate}
 
 
\paragraph{Full transition kernel.}
 
The transition kernel is the pushforward:
\[
T(U\mid s_t,a_t)
\;=\;
\int_{Y(\tilde s_t)}
\mathbf 1\!\big[\mathrm{Post}(\tilde s_t,y)\in U\big]\,
\mathbb K_{\tilde s_t}(dy),
\]
for measurable $U\subseteq \mathsf{S}$, where $\tilde s_t:=\mathrm{Pre}(s_t,a_t)$.
 
 
\paragraph{Cost Function.}
 
Fix a target set $\Phi^*\subseteq \mathcal{CS}^0_0$ of sequents that the planner aims to fully resolve in the shared library. We define the per-step cost as
\[
c(s_t,a_t) := \sum_{s\in\Phi^*} \mathbf{1}\!\big[s\notin \mathcal{FRS}^0_t\big],
\]
the number of target sequents not yet fully resolved in the shared library at time $t$.
 
\begin{remark}[Cost variants]
Under this cost function, the finite-horizon objective $J_H(\pi) = \sum_{t=0}^{H-1}\mathbb{E}[c(s_t,a_t)]$ is minimized by policies that resolve target sequents as early as possible, yielding a natural connection to weighted completion time in scheduling theory. For a pure \emph{makespan} objective (minimizing the time at which all targets are resolved), one may instead consider the stopping-time formulation
\[
\mathrm{makespan}(\pi) := \inf\!\big\{t\ge 0 : \Phi^*\subseteq \mathcal{FRS}^0_t\big\}
\]
and seek to minimize $\mathbb{E}^\pi[\mathrm{makespan}(\pi)]$. The discounted formulation with $c$ as above interpolates between these via the discount factor $\gamma$.
\end{remark}


\paragraph{Initial distribution $T^0$.}

$T^0$ is a distribution over initial states $s_0$ specifying initial private and shared libraries (satisfying $\mathcal{L}^0_0\subseteq\mathcal{L}^\alpha_0$), idle job statuses ($J^\alpha_0=\bot$ for all $\alpha$), zero cumulative-effort maps ($U^\alpha_0\equiv 0$), initial query models $Q^\alpha_0\in\mathcal{Q}$, and empty query buffers ($\Omega^\alpha_0=\bot$), supported on states satisfying all invariants.


\subsubsection{Example}\label{sec:matp-example}
\textbf{TODO REPLACE WITH A TRIANGLE THAT SHOWS PARALLEL WORK WITH STOCHASTICITY}

We work through a simple instance to illustrate the centralized MATP formulation.

\paragraph{Setup.} Let $\mathcal{N}=\{\alpha_1,\alpha_2\}$ be two agents with identical, deterministic proof kernels. The shared library $\mathcal{L}^0_0$ contains three concrete formulas $\{A,B,C\}$ (and their negations), with the single axiom $[\proves A]\in\mathcal{FRS}^0_0$. The planner's target is $\Phi^*=\{[\proves C]\}$. The dependency structure of $C$ requires two intermediate lemmas:
\[
[A\proves B]\quad\text{and}\quad [A,B\proves C],
\]
meaning that proving $C$ from axioms requires first proving $B$ from $A$, then $C$ from $A$ and $B$. All libraries are public (the PPI setting of Remark~\ref{rem:perfect-public-info}), so the POMDP reduces to an MDP. Each proof attempt takes exactly $\tau=1$ step and succeeds deterministically.

\paragraph{Optimal plan.} At $t=0$, the planner assigns $\alpha_1\mapsto (\mathrm{Prove},([A\proves B],1))$ and $\alpha_2\mapsto \bot$. Agent $\alpha_2$ is forced to idle because it cannot yet attempt $[A,B\proves C]$, as $B$ has not been proven. At $t=1$, $\alpha_1$'s proof completes, resolving $[A\proves B]$ in the (fully public) library. The planner then publishes it and assigns either agent to $(\mathrm{Prove},([A,B\proves C],1))$, which resolves $C$ at $t=2$.

\paragraph{Discussion.} The sequential dependency $A\to B\to C$ prevents any parallel speedup despite having two agents, illustrating how the precedence structure of the dependency DAG constrains the planner's scheduling freedom. This observation motivates the reduction to stochastic scheduling with precedence constraints developed in Section~\ref{sec:scheduling}.

Even in this small instance, if the libraries were private (i.e., $\alpha_1$ discovers $[A\proves B]$ but has not yet published it), the planner would need to decide \emph{when} to instruct $\alpha_1$ to publish, trading off the publication cost (one timestep) against enabling $\alpha_2$. The planner's inability to directly observe $\alpha_1$'s private library in the general (non-PPI) setting further complicates scheduling, as the planner must infer what each agent has privately proven from the history of query responses and publications.

\subsection{MATP and Scheduling Theory}\label{sec:scheduling}

We now analyze the centralized MATP problem under two simplifying restrictions that expose its combinatorial core. First, we assume the \emph{perfect public information} (PPI) setting of Remark~\ref{rem:perfect-public-info}, in which all libraries are public ($\mathcal{L}^\alpha_t = \mathcal{L}^0_t$ for all $\alpha,t$); this eliminates partial observability and reduces the POMDP to an MDP. Second, we disable conjecturing, so that the set of concrete formulas is fixed. Under these two restrictions, the centralized MATP acquires a natural interpretation as a stochastic scheduling problem with latent dependencies, allowing us to import complexity and approximation results from scheduling theory, while also identifying MATP-specific structural features that go beyond the classical scheduling setting.


\subsubsection{Non-Conjecturing PPI MATP}\label{sec:nc-ppi}

\begin{definition}[Non-Conjecturing PPI MATP]\label{def:nc-ppi-matp}
A \emph{Non-Conjecturing PPI MATP} instance is a centralized MATP instance $\mathcal{P}_{\mathrm{MATP}}$ subject to:
\begin{enumerate}
    \item \textbf{Perfect public information (PPI).} $\mathcal{L}^\alpha_t = \mathcal{L}^0_t$ for all $\alpha\in\mathcal{N},\; t\in\mathbb{N}$. Publication is vacuous, and the POMDP reduces to a fully observed MDP.
    
    \item \textbf{No conjecturing.} The conjecture action is disabled, and $\mathcal{C}_0 = \mathcal{F}$: every formula is already concrete at initialization. Thus $\mathcal{CS}_0 = \mathcal{P}_{\mathrm{fin}}(\mathcal{F})/{\sim}\;\times\;\mathcal{F}$ is the full set of concrete sequents, fixed for all time.
    
    \item \textbf{Unique dependencies.} For each sequent $s\in\mathcal{S}$, the dependency set $d(s)\subseteq\mathcal{F}$ produced upon successful proof is unique: there is a single deterministic function $\delta^*:\mathcal{S}\to\mathcal{P}_{\mathrm{fin}}(\mathcal{F})$ such that whenever any agent $\alpha$ successfully resolves $s$ (i.e., $(1,d)\sim\mathcal{K}^\alpha_{\mathrm{proof}}(\cdot\mid\mathcal{L},s,\tau)$), we have $d = \delta^*(s)$. The function $\delta^*$ is \emph{latent}: the planner does not know $\delta^*(s)$ until $s$ is resolved.
\end{enumerate}
\end{definition}

\paragraph{The latent task set.}

Since dependencies are unique and deterministic, the set of sequents that must ultimately be resolved to fully resolve the target set $\Phi^*$ is itself deterministic---though unknown to the planner.

\begin{definition}[Dependency closure and latent task set]\label{def:latent-tasks}
For a target set $\Phi^*\subseteq\mathcal{CS}_0$, define the \emph{latent task set} $\mathcal{T}^* = \mathrm{cl}(\Phi^*)$ as the smallest set satisfying:
\begin{enumerate}
    \item $\Phi^*\subseteq\mathcal{T}^*$.
    \item For every $s = [\Gamma\proves\phi]\in\mathcal{T}^*$ that is not already an axiom, and every $\psi\in\delta^*(s)$, there exists a sequent $s' = [\Gamma'\proves\psi]$ with $\Gamma'\subseteq\Gamma$ such that $s'\in\mathcal{T}^*$.
\end{enumerate}
By the acyclicity invariant on $\delta^*$, this closure is well-defined and finite (assuming $\Phi^*$ is finite and $\delta^*$ has finite image).
\end{definition}

We emphasize that $\mathcal{T}^*$ is a deterministic set, fixed by $\Phi^*$ and the latent function $\delta^*$. However, the planner does not know $\mathcal{T}^*$ at time $0$. Instead, $\mathcal{T}^*$ is revealed incrementally: when the planner resolves some $s\in\mathcal{T}^*$, the dependency set $\delta^*(s)$ becomes visible, and the planner learns which additional sequents must be resolved.

\begin{remark}[No operational precedence]\label{rem:no-operational-prec}
A key structural feature: the planner may attempt to resolve \emph{any} concrete sequent at any time. There is no operational precedence constraint requiring dependencies to be resolved before the sequent itself. The constraint is purely on the \emph{objective}: a target sequent $s\in\Phi^*$ is \emph{fully resolved} (i.e., $s\in\mathcal{FRS}_t$) only when $s$ is resolved \emph{and} all sequents in its (transitive) dependency closure are fully resolved. Thus the dependency structure constrains the \emph{completion criterion}, not the \emph{execution order}. 

This is fundamentally different from classical scheduling with precedence constraints, where predecessor jobs must complete before successor jobs may begin.
\textbf{TODO FIX No? i mean technically one can attempt to prove whatever they like whenever they like, things won't be fully resolved until the dependencies are resolved. And you find what the dependencies are as you prove things, so you can attempt to prove things that turn out to be irrelevant, but you won't get credit for them until the dependencies are resolved. so idk}
\end{remark}


\paragraph{Information structure and revelation.}

The planner's information at time $t$ consists of the full public library state $\mathcal{L}^0_t$, including all resolved sequents $\mathcal{RS}^0_t$ and their (now-revealed) dependency sets $\{\delta^*(s) : s\in\mathcal{RS}^0_t\}$. The latent information is: (i) the dependency sets $\delta^*(s)$ for unresolved sequents $s\notin\mathcal{RS}^0_t$, and (ii) the realized processing times for incomplete proof attempts.

The query model provides noisy ex-ante estimates: for any concrete sequent $s$, the planner may query agent $\alpha$ to obtain $(\hat p, \hat\tau) = q^\alpha(Q^\alpha_t, s)$, estimating the probability that $s$ is resolvable by $\alpha$ and the expected time to do so. These estimates are updated as proofs complete and the query model state $Q^\alpha_t$ is refined. This allows the planner to form beliefs about which sequents are likely to be in $\mathcal{T}^*$ and how difficult they are, enabling \emph{speculative} proof attempts on sequents that may turn out to be needed.


\subsubsection{Scheduling Interpretation}\label{sec:sched-interp}

The Non-Conjecturing PPI MATP admits a natural interpretation as a stochastic scheduling problem, but with important structural differences from the classical setting.

\paragraph{Correspondence.}

Fix a Non-Conjecturing PPI MATP instance with $m = |\mathcal{N}|$ agents and latent task set $\mathcal{T}^*$. The scheduling interpretation is:
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
\textbf{MATP} & \textbf{Scheduling} \\
\hline
Agent $\alpha\in\mathcal{N}$ & Machine $M_\alpha$ (unrelated) \\
Sequent $s\in\mathcal{T}^*$ & Job $J_s$ \\
Time-to-solution $T^\alpha_s$ & Processing time $P_{\alpha s}$ \\
Target set $\Phi^*$ & Target job set $\mathcal{J}^{\mathrm{target}}$ \\
Dependency $\psi\in\delta^*(s)$ & Spawned job upon completion of $J_s$ \\
Cumulative effort $U^\alpha_t(s)$ & Preemptive progress (machine-specific) \\
OR-redundancy & Multiple machines on same job \\
\end{tabular}
\end{center}

\paragraph{Stochastic processing times.}

For each job $J_s$ and machine $M_\alpha$, the processing time is the random variable $T^\alpha_s$ induced by the proof kernel:
\[
\Pr(T^\alpha_s \le \tau) = p^\alpha_\tau(s) := \sum_{d\in\mathcal{P}_{\mathrm{fin}}(\mathcal{F})} \mathcal{K}^\alpha_{\mathrm{proof}}\big((1,d)\mid\mathcal{L}^0_0,s,\tau\big).
\]
If $p^\alpha_\infty(s) < 1$, the processing time is infinite with positive probability, corresponding to agent $\alpha$ being unable to resolve $s$.

\paragraph{Key departures from classical scheduling.}

While the correspondence above is suggestive, the Non-Conjecturing PPI MATP differs from classical scheduling in three essential ways:

\begin{enumerate}
    \item \textbf{No operational precedence, latent dependency revelation.} The job set $\mathcal{T}^*$ is not known to the scheduler at time $0$. Only the target jobs $\Phi^*$ are known initially. When job $J_s$ completes, the dependency set $\delta^*(s)$ is revealed, and the corresponding jobs $\{J_{s'} : s'\text{ induced by }\delta^*(s)\}$ enter the scheduler's known task queue. This is a \emph{job-spawning} model: completing a job may generate new jobs. There is no restriction on the order in which jobs may be \emph{started}---the scheduler may attempt any concrete sequent at any time.
    
    \item \textbf{OR-redundancy.} Multiple agents may simultaneously attempt the same sequent. The sequent is resolved when \emph{any} agent succeeds, and the effective processing time is $T_s^{\mathrm{OR}} = \min_{\alpha\in A_s} T^\alpha_s$ for the set $A_s$ of agents assigned to $s$. This is absent in standard scheduling, where each job is processed by exactly one machine per timestep.
    
    \item \textbf{Machine-specific, non-transferable progress.} When agent $\alpha$ spends time $\tau$ on sequent $s$ without success, the cumulative effort $U^\alpha_t(s,\mathrm{proof})$ is updated and the next attempt by $\alpha$ on $s$ benefits from the accumulated budget. However, agent $\beta$'s progress on $s$ is independent: $U^\beta$ does not benefit from $\alpha$'s efforts. Progress is preemptible (an agent may pause and resume) but not transferable across agents.
    \textbf{TODO FIX: Maybe want to change def to allow for inter-machine progress transfer? Or maybe not bc private info is hard to transfer?}
\end{enumerate}


\paragraph{Revelation DAG.}

Define the \emph{revelation DAG} $G^{\mathrm{rev}} = (\mathcal{T}^*, E^{\mathrm{rev}})$ by
\[
E^{\mathrm{rev}} = \{(s, s') : s'\text{ is induced by some }\psi\in\delta^*(s)\},
\]
i.e., there is an edge from $s$ to each job spawned by $s$'s completion. A \emph{revelation chain} is a directed path $s_0\to s_1\to\cdots\to s_k$ in $G^{\mathrm{rev}}$, with the interpretation that $s_{i+1}$ becomes known to the planner only after $s_i$ is resolved. The acyclicity of $\delta^*$ ensures $G^{\mathrm{rev}}$ is a DAG. Denote its depth by $\mathrm{depth}(G^{\mathrm{rev}})$.


\subsection{Complexity and Approximability}\label{sec:complexity}

We now establish the computational complexity of the Non-Conjecturing PPI MATP and develop approximation bounds. Throughout, $m = |\mathcal{N}|$ is the number of agents, $n^* = |\mathcal{T}^*|$ is the size of the latent task set, and $n_0 = |\Phi^*|$ is the number of initial targets.


\subsubsection{Complexity Results}\label{sec:complexity-classes}


\paragraph{Embedding of classical scheduling.}

The Non-Conjecturing PPI MATP contains several classical scheduling problems as special cases, obtained by restricting the dependency structure and agent heterogeneity.

\begin{proposition}[Scheduling embeddings]\label{prop:embeddings}
The following scheduling problems embed into the Non-Conjecturing PPI MATP:
\begin{enumerate}
    \item \textbf{$P\|\, C_{\max}$ (identical parallel machines, makespan).} Take all agents identical ($\mathcal{K}^{\alpha_i}_{\mathrm{proof}} = \mathcal{K}^{\alpha_j}_{\mathrm{proof}}$), deterministic unit-success proof kernels, $\Phi^* = \mathcal{T}^*$ (no latent tasks), and $\delta^*(s) = \varnothing$ for all $s$ (trivial dependencies). Then the MATP reduces to assigning $n^*$ jobs with known deterministic processing times to $m$ identical machines to minimize makespan.
    
    \item \textbf{$R\|\, C_{\max}$ (unrelated machines, makespan).} As above but with heterogeneous proof kernels.
    
    \item \textbf{$P\mid\mathrm{prec}\mid C_{\max}$ (identical machines, precedence, makespan).} Take identical agents, $\delta^*(s) = \varnothing$ for all $s$, but introduce operational precedence by encoding it as follows: include ``sentinel'' sequents whose role is to enforce ordering. Specifically, if job $J_j$ must precede $J_k$, include a sentinel sequent $s_{jk}$ with $\delta^*(s_{jk}) = \{\phi_j\}$ (where $\phi_j$ is the consequent of $J_j$) and make $s_{jk}$ a dependency of a target that also depends on $J_k$'s consequent. The unique-dependency and latent-revelation structure then enforces the desired precedence: $s_{jk}$ cannot be marked fully resolved until $J_j$ completes, which in turn blocks full resolution of any target depending on $s_{jk}$.

    In fact, this encoding is unnecessary in the special case where $\delta^*(s)=\varnothing$ for all $s$ and all targets are independent, but becomes relevant when one considers the full dependency structure.
\end{enumerate}
\end{proposition}

The embedding of $P\|\,C_{\max}$ is the most direct and yields the following.

\begin{theorem}[NP-hardness of MATP]\label{thm:np-hard}
The decision problem for the Non-Conjecturing PPI MATP---``given an instance $\mathcal{P}$ and threshold $T$, is $\inf_\pi \mathbb{E}^\pi[C_{\max}] \le T$?''---is NP-hard, even restricted to:
\begin{itemize}
    \item identical agents,
    \item deterministic proof kernels (processing times are deterministic),
    \item trivial dependencies ($\delta^*(s) = \varnothing$ for all $s$), and
    \item single-assignment (no OR-redundancy).
\end{itemize}
More precisely, $P\|\, C_{\max}$ reduces in polynomial time to this restricted MATP. Since $P\|\, C_{\max}$ is strongly NP-hard for variable $m$ (and weakly NP-hard for $m=2$), the same holds for MATP.
\end{theorem}

\begin{proof}
Given an instance of $P\|\, C_{\max}$ with $n$ jobs of processing times $p_1,\dots,p_n$ on $m$ identical machines, construct a Non-Conjecturing PPI MATP instance with $m$ identical agents, target set $\Phi^* = \{s_1,\dots,s_n\}$, and deterministic proof kernels such that agent $\alpha$ resolves $s_j$ in exactly $p_j$ time steps (i.e., $\mathcal{K}^\alpha_{\mathrm{proof}}((1,\varnothing)\mid\mathcal{L},s_j,\tau) = \mathbf{1}[\tau \ge p_j]$). Set $\delta^*(s_j) = \varnothing$ for all $j$. Then every resolved sequent is automatically fully resolved, and the MATP makespan equals the scheduling makespan. The reduction is polynomial in the input size.
\end{proof}

\paragraph{State space and MDP complexity.}

Under PPI, the MATP is a fully observed MDP, and in principle one may apply standard MDP algorithms (value iteration, policy iteration) to compute the optimal policy. However, the state space is exponentially large in the problem parameters.

\begin{proposition}[State space exponentiality]\label{prop:state-space}
The state space $\mathsf{S}$ of the Non-Conjecturing PPI MATP MDP has cardinality at least $2^{n^*}$, where $n^* = |\mathcal{T}^*|$ is the size of the latent task set. Specifically, the component $\mathcal{RS}^0_t\subseteq\mathcal{CS}_0$ ranges over $2^{n^*}$ possible resolved-sequent sets (restricted to $\mathcal{T}^*$), and these are all reachable.
\end{proposition}

\begin{proof}
Since any subset of $\mathcal{T}^*$ can arise as the resolved set at some time $t$ (by choosing which sequents to prove and in what order), the state space has at least $2^{n^*}$ states. The additional job-descriptor and cumulative-effort components further enlarge the state space.
\end{proof}

\begin{corollary}[Exact solution complexity]\label{cor:exact-complexity}
Solving the Non-Conjecturing PPI MATP MDP exactly (via value iteration or linear programming over the state space) requires time $\mathrm{poly}(|\mathsf{S}|,|\mathsf{A}|) = 2^{\Omega(n^*)}$, and is thus intractable for large task sets unless the MDP has exploitable structure (e.g., factored representations).
\end{corollary}

\begin{remark}[PSPACE membership for the decision problem]
The decision problem ``$\inf_\pi \mathbb{E}^\pi[C_{\max}] \le T$?'' for the stochastic MATP MDP lies in PSPACE: one can enumerate policies and simulate trajectories in polynomial space, maintaining only the current state and running cost. Combined with Theorem~\ref{thm:np-hard}, the decision problem is NP-hard and in PSPACE. We conjecture it is PSPACE-complete, as the stochastic processing times and latent dependency revelation introduce planning-under-uncertainty structure analogous to stochastic reachability problems, but we do not prove this here.
\end{remark}


\subsubsection{Lower Bounds}\label{sec:lower-bounds}

We now establish information-theoretic lower bounds on the optimal makespan. These bounds are \emph{universal}: they apply to every policy, including policies that speculatively prove sequents before their necessity is confirmed.

Fix a Non-Conjecturing PPI MATP instance with $m$ agents, target set $\Phi^*$, and latent task set $\mathcal{T}^*$.

\begin{definition}[Work and serial depth]\label{def:work-depth}
For a realization $\omega$ (which determines all processing times), define:
\begin{itemize}
    \item The \emph{work} is the total processing time across all tasks in the latent task set, under an optimal single-assignment:
    \[
    W(\omega) := \sum_{s\in\mathcal{T}^*} \min_{\alpha\in\mathcal{N}} T^\alpha_s(\omega).
    \]
    \item The \emph{serial depth} is the maximum total processing time along any revelation chain:
    \[
    D(\omega) := \max_{\substack{s_0\to s_1\to\cdots\to s_k \\ \text{in } G^{\mathrm{rev}}}} \sum_{i=0}^{k} \min_{\alpha\in\mathcal{N}} T^\alpha_{s_i}(\omega).
    \]
\end{itemize}
\end{definition}

\begin{theorem}[Work and serial depth lower bounds]\label{thm:lower-bounds}
For any Non-Conjecturing PPI MATP instance and any admissible policy $\pi$:
\[
\mathbb{E}^\pi[C_{\max}] \;\ge\; \max\!\left\{\frac{\mathbb{E}[W]}{m},\;\; \mathbb{E}[D]\right\}.
\]
\end{theorem}

\begin{proof}
\textbf{Work bound.} For each realization $\omega$, every task $s\in\mathcal{T}^*$ must be resolved by some agent. Even if each task is resolved by the fastest agent (giving processing time $\min_\alpha T^\alpha_s(\omega)$), the total work is $W(\omega)$. With $m$ agents working in parallel, the makespan is at least $W(\omega)/m$. Taking expectations: $\mathbb{E}^\pi[C_{\max}] \ge \mathbb{E}[W]/m$.

\textbf{Serial depth bound.} Consider any revelation chain $s_0\to s_1\to\cdots\to s_k$ in $G^{\mathrm{rev}}$. By definition, $s_{i+1}$ is a dependency of $s_i$, and $\delta^*(s_i)$ is revealed only when $s_i$ is resolved. Thus $s_{i+1}$ is not known to be required until time $C_{s_i}$ (the completion time of $s_i$). Therefore, $s_{i+1}$ cannot be completed before time $C_{s_i} + \min_\alpha T^\alpha_{s_{i+1}}(\omega)$ at the earliest (assigning the fastest available agent immediately upon revelation). Iterating: the completion of the chain requires time at least $\sum_{i=0}^k \min_\alpha T^\alpha_{s_i}(\omega)$. Since all tasks in the chain must be resolved for $s_0$ to be fully resolved (and $s_0\in\Phi^*$ implies all its descendants are in $\mathcal{T}^*$), the makespan is at least $D(\omega)$. Taking expectations gives the bound.
\end{proof}

\begin{remark}[Tightness of the serial depth bound]\label{rem:serial-depth-tight}
The serial depth bound is tight in the following sense: if the revelation DAG has a single chain $s_0\to s_1\to\cdots\to s_k$ with all other tasks having trivial dependencies, then no policy (not even one with perfect foresight about $\delta^*$) can fully resolve $s_0$ faster than $\sum_i \min_\alpha T^\alpha_{s_i}(\omega)$, because the existence of $s_{i+1}$ is not known until $s_i$ completes.
\end{remark}

\begin{remark}[Speculation and the serial depth bound]\label{rem:speculation}
One might wonder whether a policy that \emph{speculatively} proves sequents---resolving $s_{i+1}$ \emph{before} $s_i$ completes, on the guess that $s_{i+1}$ will be needed---could beat the serial depth bound. It cannot: by assumption, $\delta^*$ is latent, and the planner does not know $s_{i+1}\in\delta^*(s_i)$ until $s_i$ is resolved. If the planner speculatively proves some sequent $s'$ that turns out not to be in $\mathcal{T}^*$, this work is wasted. If $s'$ does turn out to be needed, the speculative work is useful, but the planner still cannot resolve $s'$ before it is discovered to be in $\mathcal{T}^*$---the resolution may have already occurred speculatively, but the \emph{revelation} of $s'$ as a required task still occurs only upon completion of its parent in the revelation DAG.

More precisely: speculation can help the \emph{work bound} (by resolving tasks that will later be needed, before they are officially revealed). But speculation cannot help the \emph{serial depth bound}, because the serial depth counts the sum of processing times along a chain where each link is only discovered upon completion of the previous link---and if a speculative proof of $s_{i+1}$ was already completed before $s_i$, then $\min_\alpha T^\alpha_{s_{i+1}} = 0$ in the chain, and the serial depth adjusts accordingly.

To formalize: define the \emph{residual serial depth} under policy $\pi$ as the serial depth computed with processing times replaced by the \emph{residual} time $\max\{0, \min_\alpha T^\alpha_{s_i}(\omega) - U^\alpha_{C_{s_{i-1}}}(s_i)\}$, accounting for any speculative work already completed. The bound $C_{\max}(\omega) \ge D_{\mathrm{res}}(\omega)$ still holds, and in the worst case (no useful speculation), $D_{\mathrm{res}} = D$.
\end{remark}


\subsubsection{Approximation Algorithms}\label{sec:approx}


We now develop approximation guarantees for the Non-Conjecturing PPI MATP. The natural algorithmic paradigm is \emph{reactive scheduling}: maintain a queue of known required tasks (initialized to $\Phi^*$, expanded as dependencies are revealed), and apply a scheduling heuristic to the current queue.

\begin{definition}[Reactive list scheduling]\label{def:reactive-ls}
Fix a priority function $\sigma:\mathcal{CS}_0\to\mathbb{R}$ (lower values indicate higher priority). The \emph{Reactive List Scheduling} (RLS) policy operates as follows:
\begin{enumerate}
    \item Initialize the \emph{known task queue} $\mathcal{K}_0 := \Phi^*$.
    \item At each timestep $t$: for each idle agent $\alpha$, assign it to the highest-priority task in $\mathcal{K}_t$ that is not yet resolved and not currently being worked on by another agent (under single-assignment), or $\bot$ if no such task exists.
    \item When a task $s$ is resolved at time $t$, observe $\delta^*(s)$ and update:
    \[
    \mathcal{K}_{t+1} \leftarrow \mathcal{K}_t \;\cup\; \{s' : s'\text{ is induced by }\psi\in\delta^*(s),\; s'\notin\mathcal{RS}^0_t\}.
    \]
\end{enumerate}
\end{definition}

\begin{remark}[OR-redundant variant]
In the OR-redundant variant of RLS, step (2) allows multiple idle agents to be assigned to the same task. This is advantageous when $|\mathcal{K}_t| < m$ (fewer known tasks than agents), as otherwise agents would idle.
\end{remark}


\paragraph{Identical agents, single-assignment.}

We first analyze the simplest nontrivial case: $m$ identical agents with single-assignment.

\begin{theorem}[RLS approximation for identical agents]\label{thm:rls-approx}
Consider a Non-Conjecturing PPI MATP instance with $m$ identical agents (all proof kernels equal) and single-assignment. Let $\mathrm{RLS}$ denote the reactive list scheduling policy with an arbitrary priority function. Then for each realization $\omega$:
\[
C_{\max}^{\mathrm{RLS}}(\omega) \;\le\; \frac{W(\omega)}{m} + \left(1 - \frac{1}{m}\right) D(\omega),
\]
where $W(\omega) = \sum_{s\in\mathcal{T}^*} T_s(\omega)$ and $D(\omega)$ is the serial depth. Consequently:
\[
\mathbb{E}[C_{\max}^{\mathrm{RLS}}] \;\le\; \left(2 - \frac{1}{m}\right)\mathbb{E}[\mathrm{OPT}].
\]
\end{theorem}

\begin{proof}
Fix a realization $\omega$ and let $s^*$ be the last task to complete under $\mathrm{RLS}$. Since all tasks in $\mathcal{T}^*$ must be resolved, and $s^*$ is the last to finish, we have $C_{\max}^{\mathrm{RLS}}(\omega) = C_{s^*}(\omega)$.

Partition the interval $[0, C_{s^*}]$ into:
\begin{itemize}
    \item \textbf{Busy intervals}: timesteps at which all $m$ agents are working on some task.
    \item \textbf{Starved intervals}: timesteps at which at least one agent is idle because no known unresolved task is available (all tasks in $\mathcal{K}_t$ are either resolved or currently being processed by other agents).
\end{itemize}

\textit{Busy intervals.} During busy intervals, the $m$ agents collectively process tasks at rate $m$. The total busy time is therefore at most
\[
B(\omega) \le \frac{1}{m}\sum_{s\in\mathcal{T}^*} T_s(\omega) = \frac{W(\omega)}{m}.
\]

\textit{Starved intervals.} During a starved interval, fewer than $m$ tasks are available in $\mathcal{K}_t$ (net of those currently being processed). In particular, at least one agent is idle, and every task in $\mathcal{K}_t$ is either resolved, currently being processed, or not yet known. The idle agents are waiting for a currently-running task to complete and reveal new dependencies.

We claim the total starved time can be charged to the revelation chain ending at $s^*$. To see this, consider the sequence of tasks that ``blocked'' $s^*$ from earlier completion. Let $s^* = s^*_0$. Since $s^*$ is in $\mathcal{T}^*$, either $s^*\in\Phi^*$ (known from the start) or $s^*$ was revealed as a dependency of some earlier task $s^*_1$ that was completed during RLS. In the latter case, $s^*_1\to s^*_0$ in $G^{\mathrm{rev}}$. Repeating, we obtain a chain $s^*_\ell\to\cdots\to s^*_1\to s^*_0$ in $G^{\mathrm{rev}}$ with $s^*_\ell\in\Phi^*$.

During each starved interval, the interval ends when some running task completes. If $s^*$ was in the queue but being processed, the starved interval contributes to the processing time of $s^*$ itself. If $s^*$ was not yet in the queue, the starved interval is bounded by the remaining processing time of the task in the revelation chain whose completion will (eventually) reveal $s^*$.

In either case, the total starved time is bounded by the total processing time along the revelation chain minus the processing time of $s^*$ itself (which is counted in the final step):
\[
I(\omega) \le \sum_{i=0}^{\ell} T_{s^*_i}(\omega) - T_{s^*_0}(\omega) \le D(\omega) - T_{s^*_0}(\omega).
\]

Combining:
\begin{align*}
C_{\max}^{\mathrm{RLS}}(\omega) &= B(\omega) + I(\omega) + T_{s^*_0}(\omega) \\
&\le \frac{W(\omega)}{m} + D(\omega) - T_{s^*_0}(\omega) + T_{s^*_0}(\omega) \\
&= \frac{W(\omega)}{m} + D(\omega).
\end{align*}

Wait---this gives $W/m + D$, not $W/m + (1-1/m)D$. We refine as follows. The busy-interval bound double-counts the processing of tasks in the chain: the chain tasks contribute both to $W/m$ (via the total work) and to the chain bound. Separating the chain from the remaining work:
\[
C_{\max}^{\mathrm{RLS}}(\omega) \le \frac{W(\omega) - D_{\mathrm{chain}}(\omega)}{m} + D_{\mathrm{chain}}(\omega) = \frac{W(\omega)}{m} + \left(1-\frac{1}{m}\right)D_{\mathrm{chain}}(\omega),
\]
where $D_{\mathrm{chain}}(\omega) = \sum_{i=0}^\ell T_{s^*_i}(\omega) \le D(\omega)$.

Since $\mathrm{OPT}(\omega) \ge \max\{W(\omega)/m,\; D(\omega)\}$ (Theorem~\ref{thm:lower-bounds}):
\[
C_{\max}^{\mathrm{RLS}}(\omega) \le \frac{W(\omega)}{m} + \left(1-\frac{1}{m}\right)D(\omega) \le \left(2-\frac{1}{m}\right)\mathrm{OPT}(\omega).
\]
Taking expectations completes the proof.
\end{proof}


\begin{remark}[Comparison to Graham's theorem]
Theorem~\ref{thm:rls-approx} is the MATP analogue of Graham's classical $(2-1/m)$-approximation for $P\mid\mathrm{prec}\mid C_{\max}$. The proof structure parallels Graham's: decompose time into busy and ``starved'' intervals, bound busy time by $W/m$, and bound starved time by the longest chain. The key difference is that Graham's ``longest precedence chain'' (which bounds forced-idle time due to \emph{operational} precedence) is replaced by the serial depth (which bounds forced-idle time due to \emph{informational} precedence---the online revelation of dependencies).
\end{remark}


\paragraph{Unrelated agents, single-assignment.}

When agents are heterogeneous, the problem inherits the additional complexity of the unrelated-machines model. The following is a direct consequence of the Lenstra--Shmoys--Tardos LP rounding scheme.

\begin{proposition}[Unrelated agents, no dependencies]\label{prop:unrelated-no-dep}
For a Non-Conjecturing PPI MATP instance with $m$ unrelated agents, deterministic processing times, trivial dependencies ($\delta^*\equiv\varnothing$), and single-assignment, the problem reduces to $R\|\, C_{\max}$, which admits a $2$-approximation via LP rounding (Lenstra--Shmoys--Tardos). Moreover, $R\|\, C_{\max}$ cannot be approximated within a factor of $\frac{3}{2} - \varepsilon$ for any $\varepsilon > 0$ unless $\mathrm{P} = \mathrm{NP}$.
\end{proposition}

With nontrivial dependencies and unrelated agents, the approximation landscape is less favorable, and the interaction between agent heterogeneity, latent dependencies, and OR-redundancy remains an open direction.


\paragraph{OR-redundant reactive list scheduling.}

When the number of known tasks is smaller than the number of agents ($|\mathcal{K}_t| < m$), single-assignment RLS leaves agents idle. The OR-redundant variant assigns excess agents to tasks already being processed by other agents, reducing expected completion times via parallel attempts.

\begin{proposition}[Benefit of OR-redundancy]\label{prop:or-benefit}
Consider a Non-Conjecturing PPI MATP with $m$ identical agents, a single target sequent $\Phi^* = \{s_0\}$, and a revelation chain $s_0\to s_1\to\cdots\to s_k$ of depth $k+1$ (i.e., $\delta^*(s_i) = \{\phi_{i+1}\}$ for $i < k$ and $\delta^*(s_k) = \varnothing$), with i.i.d.\ $\mathrm{Exp}(\lambda)$ processing times across all agents and tasks. Then:
\begin{enumerate}
    \item Under single-assignment RLS: $\mathbb{E}[C_{\max}] = (k+1)/\lambda$.
    \item Under OR-redundant RLS (all $m$ agents assigned to the current task): $\mathbb{E}[C_{\max}] = (k+1)/(m\lambda)$.
\end{enumerate}
Thus OR-redundancy yields a factor-$m$ speedup on pure revelation chains.
\end{proposition}

\begin{proof}
Under single-assignment, each task is processed by one agent with $\mathrm{Exp}(\lambda)$ time. The $k+1$ tasks are processed serially (each is discovered only upon completion of the previous), giving total expected time $(k+1)/\lambda$.

Under OR-redundancy, all $m$ agents work on each task. By the memoryless property, the minimum of $m$ independent $\mathrm{Exp}(\lambda)$ variables is $\mathrm{Exp}(m\lambda)$ with mean $1/(m\lambda)$. The $k+1$ tasks still require serial processing, but each takes $1/(m\lambda)$ in expectation, giving $(k+1)/(m\lambda)$.
\end{proof}

\begin{remark}[OR-redundancy vs.\ task parallelism]\label{rem:or-vs-standard}
Proposition~\ref{prop:or-benefit} shows that OR-redundancy is maximally beneficial on pure revelation chains, where task parallelism is impossible. In contrast, when the revelation DAG is wide (many independent tasks), task parallelism dominates and OR-redundancy provides little advantage. The optimal policy must balance these two forms of parallelism based on the DAG structure and processing time distributions. Computing the optimal balance is itself NP-hard in general.
\end{remark}


\subsubsection{Summary}\label{sec:complexity-summary}

\begin{table}[ht]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{llll}
\toprule
\textbf{Restriction} & \textbf{Complexity} & \textbf{Best ratio} & \textbf{Reference} \\
\midrule
$m$ identical, $\delta^*\equiv\varnothing$ & str.\ NP-hard ($m\ge 3$) & $\frac{4}{3}-\frac{1}{3m}$ (LPT) & Graham \\[2pt]
$m$ identical, latent $\delta^*$ & str.\ NP-hard ($m\ge 3$) & $2-\frac{1}{m}$ (RLS) & Thm.~\ref{thm:rls-approx} \\[2pt]
$m$ unrelated, $\delta^*\equiv\varnothing$ & str.\ NP-hard & $2$ (LP round.) & LST \\[2pt]
$m$ unrelated, latent $\delta^*$ & str.\ NP-hard & open & --- \\[2pt]
Any, + OR-redundancy & $\le$ std.\ complexity & $\le$ std.\ ratio & Rem.~\ref{rem:or-vs-standard} \\
\bottomrule
\end{tabular}
\caption{Complexity and approximation landscape for Non-Conjecturing PPI MATP. ``Latent $\delta^*$'' indicates the general case with unique but unknown dependencies. ``LPT'' is Longest Processing Time first; ``RLS'' is Reactive List Scheduling (Thm.~\protect\ref{thm:rls-approx}); ``LST'' is Lenstra--Shmoys--Tardos LP rounding; ``str.'' abbreviates ``strongly.'' All approximation ratios are with respect to the expected makespan objective.}
\label{tab:complexity}
\end{table}


\textbf{NOTE THAT WE DON'T REALLY TOUCH CONJECTURING HERE, AS IT IS ReALLY HARD TO ANALYZE? CAN WE GET ANY BOUNDS ON CONJECTURING with PPI? Even just weak to show its much harder?}

\begin{remark}[Limitations and forward pointers]\label{rem:limitations}
The Non-Conjecturing PPI MATP captures the scheduling-theoretic core of centralized multi-agent theorem proving, but several features of the full MATP lie beyond this model:
\begin{enumerate}
    \item \textbf{Conjecturing.} Enabling conjecturing ($\mathcal{C}_0 \subsetneq \mathcal{F}$) introduces an online component: new sequents are generated during execution, and the planner must decide when to invest in exploration (conjecturing new lemmas) versus exploitation (proving known tasks). The problem becomes an instance of stochastic online scheduling with endogenous job generation.
    
    \item \textbf{Private information.} Dropping the PPI assumption restores the POMDP structure. The planner can no longer observe agents' private proof progress, introducing an information-elicitation problem fundamentally different from the combinatorial scheduling problem studied here. This motivates the mechanism-design analysis of Section~\ref{sec:limits-centralization}.
    
    \item \textbf{Library-state-dependent proof kernels.} If the proof kernel $\mathcal{K}^\alpha_{\mathrm{proof}}$ depends on the current library state (e.g., newly proven lemmas accelerate subsequent proofs), processing times become state-dependent and correlated, breaking the independence assumptions underlying the scheduling model. This induces positive externalities between proof tasks that further complicate the optimization.
    \textbf{I DOn't think library state dependency is necessarily part of MATP, as we really only care about sequents, which are all more or less independent. However, i suppose as a job is complete when the sequent is fully resolved, which depends on how long its taks for its dependencies to be resolved, there is some indirect dependence on library state? idk}
\end{enumerate}
\end{remark}

\textbf{ALSO WE SHOULD PROBABLY PROVE THAT AN OPTIMAL POLICY FOR A FULL MATP WITHOUT CONJECTURING IS ALSO OPTIMAL FOR THE NON-CONJECTURING PPI VERSION}
\subsubsection{Limits of Centralization}\label{sec:limits-centralization}

The centralized MATP problem provides a strong baseline: a single planner allocates theorem-proving effort across agents so as to optimize a global objective under partial observability. However, this formulation presumes a level of centralized control that is unrealistic in the motivating multiagent setting. In practice, each prover controls its own compute, possesses rich local search state that is not fully communicable, and decides whether or not to share intermediate results. Once these assumptions are relaxed, the problem is no longer one of single-agent control but of strategic interaction under partial information. This motivates a game-theoretic formulation. 
