\chapter{ImProver 2}\label{chap:improver2}

\setlength{\parindent}{1em}

We present ImProver 2, a training pipeline for proof optimization towards a specific metric. Its core purpose is to bootstrap the capabilities of small language models (\textit{SLM}s) through repeated applications of training, generation, and evaluation/filtering of generated data for use in further iterations. To do so, it utilizes three core components: a reinforcement learning-based training stage (described in \ref{sec:irpo}), a replay buffer to balance old and newly generated data (\ref{sec:training_replay}), and multiple sources of neurosymbolic augmentation to boost generation capabilities (\ref{sec:generation}). Each are described in detail below. 


\subsection{Overview}
\label{sec:improver_spec}

Given an un-modified base language model $G_0$, at the $t$-th iteration ImProver 2's core loop aims to train the model $G_{t+1}$ from $G_t$ as follows (also depicted in Figure \ref{fig:flow_pipeline}):
\begin{enumerate}
    \item For some budget hyperparameter $n \in \mathbb{N}$, generate $n$ candidate proofs per problem in the training set using the current model $G_{t}$ (\ref{sec:generation}), providing the model with neurosymbolic augmentation (\ref{sec:ns_aug}) and a description of the metric to assist in generation. These new potential proofs form a new dataset (denoted $\mathcal{D}^{(t)}_{\mathrm{nr}}$). 

    \item The previous iteration's dataset, $\mathcal{D}^{(t-1)}_{\mathrm{re}}$ (our \textit{replay buffer}), is interleaved with $\mathcal{D}^{(t)}_{\mathrm{nr}}$ to get $\mathcal{D}^{(t)}_{\mathrm{re}}$ (see \ref{sec:training_replay}). 

    \item The model $G_t$ is trained to obtain $G_{t+1}$ (\ref{sec:irpo}). Our reinforcement learning policy of choice, known as Iterative Reasoning Preference Optimization or IRPO \cite{IRPO}, relies on preference pairs of desirable/undesirable proofs, so $\mathcal{D}^{(t)}_{\mathrm{re}}$ is filtered to remove low-quality solutions and pairs are created to form $\mathcal{D}^{(t)}_{\mathrm{IRPO}}$ (again described in \ref{sec:training_replay}). 

    \item $G_{t+1}$ is evaluated with $n$ samples on the test set, again similarly to \ref{sec:generation}. The improvement in metric score of the new proofs over the originals is calculated. 
\end{enumerate}

The loop is repeated until convergence of the average improvement score, or exhaustion of the compute budget. 

% JA: "P = {(c, x, y_0}} makes it look like P is a singleton. I think you mean to say something like: "Given a fixed problems set $\mathcal{P} \subseteq \mathcal{C} \times \mathcal{X} \times \mathcal{Y}$ where, for each $(c, x, y_0)$ in $P$, $x$ is a proof of $y_0$ in $c$" (or maybe $y_0$ is only a candidate proof, i.e. a purported proof, not necessarily validated by Lean's kernel?
% (Tate) I think you're right to point out this notation as confusing, I've re-done it to use a more intuitive description at the cost of some of the formalism. 



% Given a set of problems $\mathcal{P}$, a base language model $G_{0}$, and an optimization metric $\mu$, we aim to iteratively train models $G_i$ in order to maximize the mean improvement score of generated optimizations of input proofs taken from $\mathcal{P}$.


% More explicitly, given a fixed problems set $\mathcal{P} = \{(c,x,y_0)\} \subseteq \mathcal{C} \times \mathcal{X} \times \mathcal{Y}$ with $y_0 \in \mathcal{F}(c,x)$, we define a train/test partition of $\mathcal{P}_{\mathrm{train}} \cup \mathcal{P}_{\mathrm{test}} = \mathcal{P}$. Then starting with a base model $G_0$, at iteration $t$ we train a model $G_{t+1}$ from $G_t$ as follows:
% \begin{enumerate}

% % JA: In the next item, I replaced an em-dash "--" with a comma. A few times, you have used the em-dash where a comma is more appropriate.

% \item Generate $n$ candidate proofs per train problem $T \in \mathcal{P}_{\mathrm{train}}$ with the current model $G_{t}(\cdot\mid \Psi,\mu)$ (note the neurosymbolic augmentation via $\Psi$), forming the (no-replay) dataset: $\mathcal{D}^{(t)}_{\mathrm{nr}}$.
% \item Then, we collect the previous iteration's dataset $\mathcal{D}_{\text{re}}^{(t-1)}$ and use it to form a replay buffer, which we integrate into $\mathcal{D}_\text{nr}^{(t)}$ to get $\mathcal{D}_{\text{re}}^{(t)}$. We also postprocess this new dataset with several filtering steps to remove low-quality solutions, thereby forming the filtered dataset $\mathcal{D}^{(t)}_{\mathrm{fil}}$. We use this to construct preference pairs for our IRPO training: $\mathcal{D}^{(t)}_{\mathrm{IRPO}}$.
% \item With $G_{t}$ as the base model, train on $\mathcal{D}^{(t)}_{\mathrm{IRPO}}$ with IRPO to get $G_{t+1}$.
% \item Evaluate $G_{t+1}$ on $\mathcal{P}_{\mathrm{test}}$ with $n$ samples to get $\mathcal{D}^{(t+1)}_{\mathrm{test}}$.%=\{(c_T,x_T,y_{T,0},\{(y_{T,i},\widehat{s}_{T,i})\}_{i=1}^{n})\}_{T \in \mathcal{P}_{\mathrm{test}}}\]
% Then compute the mean improvement score $\overline{S_\mu}(\mathcal{D}^{(t+1)}_{\mathrm{test}})$. 


% % and record the mean improvement score of $G_{t+1}$ (and optionally $G_{t}^{(\mathrm{wSFT})}$ as well) on $\mathcal{P}_{\mathrm{test}}$ via best-of-$n$ (§\ref{sec:generator}).
% \end{enumerate}
% This loop is repeated until convergence of $\overline{S_\mu}(\mathcal{D}^{(t+1)}_{\mathrm{test}})$ or exhaustion of compute budget. An end-to-end depiction of this loop is given in Figure \ref{fig:flow_pipeline}.
% % Optionally, we either continue from $\theta_t$ or restart from the original base and re-train to $\theta_{t+1}$; both choices perform similarly on early iterations, with the latter scaling better (ablation in §\ref{experiments}).


\input{figures/flowchart.tex}
\subsection{Neurosymbolic Augmentation}
\label{sec:generation}

A central feature of the ImProver 2 \ pipeline is its ability to generate its own training data for future iterations. This is performed by running inference on the current model $G^{(t)}$, providing it with the theorem statement and the original proof, and requesting an improved version; this is repeated $n$ times per theorem. Additionally, the model's generation capabilities are augmented with information from the proof environment as described in the following sections.

% \subsubsection{Neurosymbolic Augmentation}

\label{sec:ns_aug}
Namly, theseformal proof environments provide substantial opportunities to obtain relevant information about a proof. We prompt our language model not only with serialized information on the original theorem statement $x$, proof $y_0$, and metric $\mu$, but also with additional neurosymbolic context that exposes structure and dependencies at both a formal and informal level. This augmentation comes from three sources: a context slice to find relevant lemmas or definitions, goal-state traces to highlight the exact effect of each tactic on the progress of the proof, and auto-informalization to provide a higher-level natural language description of the proof in question. Each of these sources is described in detail below. 


% Explicitly, this augmentation -- as modelled by the map $\Psi$ -- extracts and serializes three additional channels of information from the raw input triplet $I=(c,x,y_0)$:
% \[
% \Psi:\ I\ \mapsto\ \Big(I,\underbrace{\Psi_{\text{ctx}}(I)}_{\text{context slice}},\ \underbrace{\Psi_{\text{cos}}(I)}_{\text{goal-state traces}},\ \underbrace{\Psi_{\text{inf}}(I)}_{\text{auto-informalization}}\Big).
% \]
% The generator (LLM) sees only the serialization of $(\Psi,\mu)$, so its conditional distribution effectively factors through these channels. Now, we describe each channel in turn.

\subsubsection{Context}
\label{sec:context_extraction}
When working with proofs in high-dependency environments, it is likely that a given proof $y_0$ relies on many lemmas, definitions, and other formal objects defined in the context $c$. We aim to extract and serialize a minimal set of these objects to better inform our generation process.

To facilitate extraction of relevant context from the proof environment, we consider the Lean~4 concrete syntax tree (CST) and abstract syntax tree (AST). The CST preserves surface tokens and references locations in source code; the AST resolves names, binds variables, and canonicalizes declarations (e.g., definition/theorem/structure/etc. nodes). Context extraction uses both:
\begin{itemize}
    \item \textbf{CST view (textual reachability):} harvest all surface identifiers and their source spans that occur in $x$ and in the proof text of $y_0$.
    \item \textbf{AST view (semantic reachability):} resolve those identifiers to fully qualified symbols under the language's environment (imports, namespaces, instances, modules); record declaration categories (e.g., definition, theorem) and provenance (module/package).
\end{itemize}

\paragraph{Graphs and the slice.}
From the AST we build two standard graphs: (i) an import DAG over files/modules, and (ii) an entity graph whose vertices are declarations (constants, lemmas, definitions) with edges for semantic references (uses in types/bodies). Let $\mathsf{touch}(x,y_0)$ be the set of AST nodes directly referenced by the CST identifiers collected from $x$ and $y_0$ (optionally augmented by a dynamic trace of proof states, if available). The context slice is then the subgraph
\[S(c,x,y_0) = \mathrm{Reach}\big(G_{\mathrm{ent}},\,\mathsf{touch}(x,y_0)\big),\]
optionally restricted by the import DAG to a budgeted neighborhood. We then serialize each element of $S$ with metadata describing its object type (e.g.\ lemma, definition), along with a stable snippet of its source content.

Because selection is driven by AST resolution, $\Psi_{\text{ctx}}$ is invariant to superficial edits (whitespace, formatting) and robust to local refactors ($\alpha$-renaming within a module). The slice size grows with the reachable subgraph, not with raw file size, yielding a compact, minimal, and deterministic bundle of proof context that is read-only with respect to the program state.

\subsubsection{Chain-of-States (CoS)}
Chain-of-states prompting \cite{ImProver} provides a language model with the explicit state and remaining goals of a proof following the application of each tactic/step, providing richer information about the proof's structure than is normally available. We utilize Lean's \texttt{InfoTree} structures to obtain and serialize these states, and interleave them into the original proof as comments.

More formally, given a verified proof $y_0$, let its tactic steps be indexed by $i=1,\dots,T$. From the \texttt{InfoTree} we obtain, for each step, the pair $(\texttt{goalsBefore}_i,\texttt{goalsAfter}_i)$ along with a pretty-printed tactic $\tau_i$. We then define
\[
\Psi_{\text{cos}}(y_0)\ =\ \big\langle\,(\texttt{goalsBefore}_i,\ \tau_i,\ \texttt{goalsAfter}_i)\,\big\rangle_{i=1}^T,
\]
and serialize each triple as a short snippet in which the goal lists are pretty-printed into comments adjacent to $\tau_i$. This turns hidden kernel states into explicit cues the model can condition on.

\subsubsection{Auto-informalization}
Utilizing natural language to guide the generation of formal proofs has been shown to significantly increase the capabilities of LLM-based systems on formal mathematical tasks \cite{DraftSketchProve}. We therefore expose natural-language sketches of each target proof to the model, providing a fuzzy layer of abstraction that captures the ``meaning'' of formal items while being robust to syntactic variation and surface-level noise. 

More concretely, we prompt a language model using the proof's chain-of-states information as described above to translate a Lean proof into natural language by explaining the effect of each tactic on the proof state and providing this to the proof optimizer model. We emphasize that this informalization is a secondary channel for the generator and serves simply as an additional representation of the target theorem; outputs are still prompted to be generated formally, and correctness is still judged formally.


\subsection{Training}
\label{sec:training}

% JA: I changed "firstly" to "first".

We build upon the Iterative Reasoning Preference Optimization algorithm \cite{IRPO} to improve proof optimization capabilities during each round of training. However, our architecture differs from the original implementation in several ways. First, we use the improvement in score (i.e. $\mu(s, x, y_{1}) > \mu(s, x, y_{2})$) in addition to binary correctness to rank candidates and form preference pairs, creating a denser reward signal. We furthermore maintain a replay buffer that mixes newly-generated solutions with old data to enable stable improvement over many training rounds. 
Finally, unlike the original IRPO construction, where training is done on both the reasoning trace as well as the final output, we mask the reasoning trace during training and train solely on the final output.

\subsubsection{Datasets and Replay Buffer}
\label{sec:training_replay}
Indiscriminate consumption of self-generated training data has been shown to harm performance in language models, collapsing the tails of their distribution \cite{modelCollapse}. To avoid this issue and enable self-improvement, we introduce a novel replay buffer design that filters new samples, combines them with existing training data, and sorts them based on the degree of their improvement for use during training.

Given the problem set $\mathcal{P} = \mathcal{P}_{\mathrm{train}} \cup \mathcal{P}_{\mathrm{test}}$ and the current iteration-$t$ model $G_t$, we run generation as described in \S\ref{sec:generation} to obtain $n$ candidate proofs per problem in $\mathcal{P}_{\mathrm{train}}$. This yields the raw (no-replay) dataset
\begin{align*}
    \mathcal{D}^{(t)}_{\mathrm{nr}}
    &= \{(c_T,x_T,y_{T,0},\{(y_{T,i},\widehat{s}_{T,i})\}_{i=1}^{n})\}_{T \in \mathcal{P}_{\mathrm{train}}} \\
    &= \{(c_T,x_T,y_{T,0},\mathcal{Y}_T^{(t)})\}_{T \in \mathcal{P}_{\mathrm{train}}}
\end{align*}
where $\widehat{s}_{T,i} = S_\mu(y_{T,i},y_{T,0}\mid c_T,x_T)$ is the improvement score of candidate $y_{T,i}$ over the baseline $y_{T,0}$.

We then construct the replay dataset $\mathcal{D}^{(t)}_{\mathrm{re}}$ by post-processing $\mathcal{D}^{(t)}_{\mathrm{nr}}$ together with the previous iteration's post-processed dataset $\mathcal{D}^{(t-1)}_{\mathrm{re}}$. The procedure is parameterized by a target replay proportion $\rho\in[0,1]$, a replay mode $\texttt{mode}\in\{\texttt{mark},\texttt{join},\texttt{replace}\}$, an improvement-rate cap $\pi_{\max}\in[0,1]$, and a minimum gap $\gamma \in [0,1]$.

\paragraph{Improvement rate and eligibility.}
For problem $T$, we define its improvement rate at iteration $j$ by
\[
\pi_T^{(j)}=\frac{1}{n}\sum_{i=1}^n \mathbb{I}\!\left[\text{v}(c_T,x_T,y_{T,i}^{(j)})=1\ \wedge\ \widehat{s}_{T,i}^{(j)}>0\right].
\]
A problem is \emph{replay-eligible} at iteration $t$ iff it had at least one improved, compiling solution in some previous iteration $j<t$. We maintain a reservoir $\mathcal{E}$ of replay-eligible problems with preference for ``easy'' items (largest $\pi_T^{(j)}$ across $j<t$).

\paragraph{Replay buffer construction.}
We build the post-replay dataset in two main steps.

\begin{enumerate}
    \item \textit{Mark.} First, mark problems in $\mathcal{D}^{(t)}_{\mathrm{nr}}$ as \textsc{replay} or \textsc{frontier} so that the fraction of replay equals $\rho$:
    \begin{enumerate}
        \item Start with the subset of problems that are replay-eligible; if their fraction exceeds $\rho$, downsample them by removing the highest-$\pi_T$ items (i.e.\ the ``easiest'') until the replay fraction is $\rho$ (capped at $\pi_{\max}$).
        \item If their fraction is below $\rho$, replace uniformly-chosen frontier items by items sampled from the reservoir $\mathcal{E}$ (taken from $\mathcal{D}^{(t-1)}_{\mathrm{re}}$) until the replay fraction reaches $\rho$ as best as possible. This keeps dataset size fixed while achieving the target mix.
    \end{enumerate}
    After this step, every item is tagged as \textsc{replay} or \textsc{frontier} and the target mix is satisfied.

    \item \textit{Merge.} For each item marked \textsc{replay} with key $(c_T,x_T,y_{T,0})$, find its counterpart in $\mathcal{D}^{(t-1)}_{\mathrm{re}}$, say with candidate (multi)set $\mathcal{\widetilde{Y}}^{(t-1)}_T$. Then we case on \texttt{mode}:
    \begin{itemize}
        \item \texttt{join}: set $\mathcal{\widetilde{Y}}^{(t)}_T = \mathcal{Y}^{(t)}_T \cup \mathcal{\widetilde{Y}}^{(t-1)}_T$, deduplicated by normalized proof text. This increases candidate diversity for IRPO.
        \item \texttt{replace}: set $\mathcal{\widetilde{Y}}^{(t)}_T = \mathcal{\widetilde{Y}}^{(t-1)}_T$, overwriting the current candidates with the previous iteration's.
        \item \texttt{mark}: skip this step (no candidate-level splice). In other words, set $\mathcal{\widetilde{Y}}^{(t)}_T = \mathcal{Y}^{(t)}_T$.
    \end{itemize}
\end{enumerate}

\noindent With this, we obtain the replay dataset
\[
\mathcal{D}^{(t)}_{\mathrm{re}} = \Big\{\,\big(c_T,x_T,y_{T,0},\ \mathcal{\widetilde{Y}}^{(t)}_T\big)\ \Big|\ T\in\mathcal{P}_{\mathrm{train}}\Big\}.
\]

\paragraph{Filtering.}
Finally, we post-process $\mathcal{D}^{(t)}_{\mathrm{re}}$ by filtering out low-quality candidates and separating the samples into winner ($W$) and loser ($L$) sets. Specifically, we filter the high-improvement-rate problems (those which were sufficiently ``easy'' for the model to improve on many candidates) by removing problems $T$ with $\pi_T^{(t)} > \pi_{\max}$. That is, we define
\[
\widetilde{\mathcal{P}}^{(t)} = \mathcal{P}_{\mathrm{train}} \setminus \{T \mid \pi_T^{(t)} > \pi_{\max}\}.
\]
Then, for each problem $T$ in $\widetilde{\mathcal{P}}^{(t)}$, we partition $\mathcal{\widetilde{Y}}^{(t)}_T = W_T^{(t)} \cup L_T^{(t)}$ with
\begin{itemize}
    \item $W_T^{(t)} = \{y \in \mathcal{\widetilde{Y}}^{(t)}_T \mid \text{v}(c_T,x_T,y) = 1 \land \widehat{s}_{T,y} > \delta^{(t)}\}$, where $\delta^{(t)}$ is the $\gamma$-th percentile of all scores $\{\widehat{s}_{T,i} : T \in \widetilde{\mathcal{P}}^{(t)},\ y_i \in \mathcal{\widetilde{Y}}^{(t)}_T\}$.
    \item $L_T^{(t)} = \mathcal{\widetilde{Y}}^{(t)}_T \setminus W_T^{(t)}$.
\end{itemize}
In other words, winners are those candidates that compile and have an improvement score above the $\gamma$-th percentile threshold across all samples in the dataset, while losers are the rest. The finalized filtered post-replay dataset is
\[
\mathcal{D}_{\text{fil}}^{(t)} = \Big\{(c_T,x_T,y_{T,0},W_T^{(t)},L_T^{(t)}) \mid T \in \widetilde{\mathcal{P}}^{(t)}\Big\}.
\]


\paragraph{IRPO Dataset}
As IRPO is a preference-based training method, $\mathcal{D}_{\text{fil}}^{(t)}$ is partitioned into a set of preference pairs. By mixing and matching many potential proofs of a single theorem, we glean a higher quantity of data points per theorem than would be available through group-based RL policies such as GRPO \cite{GRPO}, compensating for the limited amount of publicly available high-quality Lean training data. Additionally, by creating pairs of compiling/non-compiling examples as well as better/worse metric score, we balance the two parallel goals of learning to write correct Lean syntax and actually making improvements to proofs; this balance is dictated by the hyperparameters $W$ and $L$ in Algorithm~\ref{alg:pref_pair_creation}.

Concretely, this process is parameterized by $W,L \in \mathbb{N}$, which control the number of winners and losers sampled per problem. For a given problem $T$ in $\widetilde{\mathcal{P}}^{(t)}$, we first deduplicate $W_T^{(t)}$ by equal scores $\widehat{s}_{T,i} = \widehat{s}_{T,j}$ and deduplicate $L_T^{(t)}$ by string equality. We then form two families of preference pairs:
\[
\mathfrak{P}^{\ell\to w}_T=\{(b,g): b\in L_T^{(t)},\ g\in W_T^{(t)}\}, \qquad
\mathfrak{P}^{w\to w}_T=\{(g',g): g,g'\in W_T^{(t)},\ \widehat{s}_{T,g}>\widehat{s}_{T,g'}\}.
\]
The IRPO training dataset is then
\[\mathcal{D}_{\text{IRPO}}^{(t)} = \bigcup_{T \in \widetilde{\mathcal{P}}^{(t)}} (\mathfrak{P}^{\ell\to w}_T \cup \mathfrak{P}^{w\to w}_T),\]
which can be reinterpreted elementwise as a collection of tuples $(c_T,x_T,y_{T,0},y_{T,\ell},y_{T,w})$.


\begin{algorithm}[tb]
  \caption{Preference Pair Creation}
  \label{alg:pref_pair_creation}
  \begin{algorithmic}
    \STATE {\bfseries Input:} filtered dataset $\mathcal{D}_{\text{fil}}^{(t)}$, hyperparameters $W \in \mathbb{N}$ and $L \in \mathbb{N}$
    \STATE Initialize $\mathcal{D}_{\text{IRPO}}^{(t)} = [ \ ]$
    \FOR{$T$ {in} $\mathcal{D}_{\text{fil}}^{(t)}$}
        \STATE De-duplicate $W_T^{(t)}$ by total improvement score
        \STATE De-duplicate $L_T^{(t)}$ by string equality
        \STATE Discard or duplicate elements uniformly at random until $|W_T^{(t)}|=W$ and $|L_T^{(t)}|=L$
        \FOR{$w_1$ in $W_T^{(t)}$}
            \FOR{$w_2$ in $W_T^{(t)}$}
                \IF{$\mu(c, x, w_1) > \mu(c, x, w_2)$}
                    \STATE $\mathcal{D}_{\text{IRPO}}^{(t)} := (w_1, w_2) \texttt{::} \mathcal{D}_{\text{IRPO}}^{(t)}$
                \ENDIF
            \ENDFOR
        \ENDFOR

        \FOR{$w$ in $W_T^{(t)}$}
            \FOR{$l$ in $L_T^{(t)}$}
                \STATE $\mathcal{D}_{\text{IRPO}}^{(t)} := (w, l) \texttt{::} \mathcal{D}_{\text{IRPO}}^{(t)}$
            \ENDFOR
        \ENDFOR
    \ENDFOR
  \end{algorithmic}
\end{algorithm}









\subsubsection{IRPO (Iterative Reasoning Preference Optimization)}
\label{sec:irpo}


With this dataset $\mathcal{D}_{\text{IRPO}}^{(t)}$, we calculate the IRPO loss on each item $T$ as the (weighted) sum of the DPO loss of the preference pair and the negative log-likelihood (NLL) loss over the winner:
\begin{align*}
\mathcal{L}_{\text{IRPO}}(T) =& \mathcal{L}_{\text{DPO}}(y_{T,\ell},y_{T,w} \mid \Psi(c_T,x_T,y_{T,0}),\mu) \\&+ \alpha\mathcal{L}_{\text{NLL}}(y_{T,\ell},y_{T,w} \mid \Psi(c_T,x_T,y_{T,0}),\mu)
\end{align*}

In the above, we represent the relevant neurosymbolic augmentation with the function $\Psi$. After training with respect to this objective for one epoch, we obtain $G_{t+1}$. 
