\chapter{Conclusion}\label{chap:conclusion}

\setlength{\parindent}{1em}

This thesis has presented two systems for automated proof optimization in Lean~4, developed as a natural two-act progression from an agentic prompting approach to a specialized trained model.

\subsection{Summary of Contributions}

ImProver introduced the proof optimization task and demonstrated its feasibility using agentic prompting around a frontier language model. Its contributions span the formulation and the machinery needed to make optimization practical: it formalizes proof optimization as a task with a flexible metric framework, introduces Chain-of-States prompting as a mechanism for annotating intermediate proof states as comments in the generation context, and adds error correction and refinement loops that iteratively rewrite generated proofs. Retrieval-augmented generation over the Mathlib library and the \textit{Theorem Proving in Lean} handbook supplies the model with relevant examples and API documentation. Together these components show that proof optimization strictly generalizes neural theorem proving, and that agentic prompting alone is already capable of producing meaningful gains on length, declarativity, and mixed objectives.

ImProver~2 addresses the cost and scalability limitations of the first system by replacing the frontier-model agent with a small, specialized model trained to optimize proofs directly. Its contributions are both methodological and empirical: an iterative self-improvement pipeline built on iterative reasoning preference optimization (IRPO), stabilized by a novel replay buffer that prevents model collapse across rounds; a neurosymbolic augmentation strategy (context extraction, Chain-of-States, auto-informalization) that consistently improves performance across model sizes; and two novel metrics, modularity and dependencies, that target the structural quality and external-lemma footprint of a proof rather than merely its length. The resulting 7B-parameter model outperforms models orders of magnitude larger on structural metrics and matches frontier models on dependencies.

% \subsection{Key Insights}

% The first insight is that formal structure exposure is the key ingredient for proof optimization. Both systems derive their performance from exposing the structure of the Lean proof environment to the model. In ImProver this is done at inference time via Chain-of-States prompting: the model sees intermediate proof states as it generates. In ImProver~2 it is done at training time via neurosymbolic augmentation: the model learns from filtered, augmented preference pairs. The same underlying principle --- giving the model access to formal symbolic data about the proof --- drives gains across both paradigms.

% A second insight is that agentic prompting, while effective, is expensive. ImProver shows that frontier LLMs with the right scaffold can perform sophisticated proof optimization, and on our evaluated datasets it produces proofs that are substantially shorter or more structured than the originals. But each theorem can require many LLM calls with document retrieval, which makes library-scale deployment prohibitively costly. This observation is what motivates moving from prompting to training.

% A third insight is that, for structural metrics, specialization matters more than scale. ImProver~2 demonstrates that a properly trained 7B model outperforms 671B models on modularity and dependencies. This suggests that structural proof optimization is a specialization problem as much as a scale problem: the bottleneck is not reasoning capacity but whether the model has been adapted to the formal structure of the task.

% Finally, reward shaping in this domain requires careful guard rails. Both systems encountered the risk of degenerate solutions: proofs that score highly on a metric without achieving its intent. ImProver mitigates this by guiding the model with human-written examples; ImProver~2 uses formal heuristics (duplicate detection, wrapper detection, nontriviality filters) to define the modularity metric in a way that is robust to reward hacking. The design of robust metrics for formal proof quality remains an open challenge.

% \subsection{Direct Comparison}

% The two systems are complementary rather than competing. On the length metric, ImProver achieves a mean improvement score of $0.355$ on miniCTX-v2 compared to ImProver~2's $0.330$. This edge is consistent with ImProver's use of an agentic strategy backed by a frontier model and explicit multi-shot example retrieval for each metric. The gap is modest, however, and ImProver's per-theorem cost is orders of magnitude higher than that of the trained small model.

% On the structural metrics ImProver~2 is decisively better. For modularity it achieves $0.143$ against ImProver's $0.088$, a $63\%$ improvement; for dependencies the gap is even larger, $0.206$ versus $0.047$. These results indicate that iterative structural supervision trains refactoring capabilities that agentic prompting alone does not reliably acquire: a model that has seen thousands of preference-ranked examples of modular rewrites internalizes what modular structure looks like in a way that in-context examples cannot capture.

% The neurosymbolic scaffold (context extraction, CoS, auto-informalization) is beneficial across both systems. Applied to a baseline DeepSeek-R1 7B model, it nearly doubles the length improvement score even before any preference optimization training, and applied to frontier models it yields similarly large gains. This suggests that the scaffold would also benefit ImProver's GPT-4o base --- a direction not explored in this thesis.

\subsection{Limitations}
\label{sec:limitations-future}

Both systems have limitations that scope the claims made above and that reviewers of the underlying papers have surfaced.

ImProver is limited by its per-theorem cost (many GPT-4o calls with document retrieval) and its dependence on a closed-source frontier model, which together make library-scale deployment impractical at current API pricing and harm reproducibility. The evaluation datasets are also small ($72 + 26 + 43$ theorems), which bounds the statistical strength of the reported gains.

ImProver~2 has its own distinct set of limitations. The modularity and dependency metrics are structural proxies, not optimized for human-maintainer preferences. The system is also a single-step generator with no internal refinement loop, so when a rewrite fails to compile or regresses on the metric the model has no recovery mechanism analogous to ImProver's error-correction pass. However, this was a concious design choice in ImProver2, as we intend to focus on studying the dynamics of bootstrapped training before the additional complexity of an agentic system like ImProver.

\subsection{Future Directions}

The limitations above motivate a cluster of follow-ups. Most directly, the single-step generation in ImProver~2 can be replaced with a trained iterative-revision policy that internalizes ImProver's refinement loop into the weights, so that the model itself decides when to stop, when to revise, and how to incorporate compiler feedback. On the metric side, the structural proxies can be grounded against human preference through a small, controlled study in which formal mathematicians compare optimized rewrites against their originals; such a study is also a natural testbed for richer, more subjective metrics such as LLM-as-judge readability scoring.


Scaling out from per-theorem optimization to library-level optimization is the largest remaining challenge. Maintaining consistency across inter-dependent rewrites --- so that a modular refactor of one theorem does not break downstream proofs that referenced its intermediate names --- is both an engineering problem and a research problem, and it is the regime in which the practical value of automated proof optimization is likely to be largest. Finally, a core motivation for this line of work is improving the training data on which neural theorem provers are bootstrapped; empirically showing that provers trained on ImProver- or ImProver~2-optimized corpora outperform those trained on unoptimized data would confirm the training-data-quality hypothesis that underlies both systems.

\subsection{Closing Remarks}

Formal mathematics libraries are growing faster than human maintainers can curate them. This thesis shows that automated proof optimization --- rewriting verified proofs to be shorter, more modular, or less dependent on external lemmas --- is a feasible and learnable task. The central insight is simple but powerful: giving a language model access to the formal structure of the proof environment, whether at inference time through Chain-of-States prompting or at training time through neurosymbolic augmentation, enables far more substantive proof transformations than prompting alone.

ImProver and ImProver~2 together establish proof optimization as a well-defined task with practical applications to library maintenance, training data quality, and the broader agenda of making formal mathematics more accessible to both humans and machines. We hope these systems serve as a foundation for further work on AI-assisted formal mathematics.
